# Fisher's exact test {#fisher}

```{block2, type='wikipedia'}
<https://en.wikipedia.org/wiki/Fisher%27s_exact_test>
```

```{block2, type='youtube'}
<https://www.youtube.com/watch?v=a7ESQKI7nao&list=PL018X5Hlr4RkgE65Pg93TFY-32KCVpW84&index=16>
```

```{block2, type='youtube'}
<https://www.youtube.com/watch?v=O9LUvIFSlRo&index=17&list=PL018X5Hlr4RkgE65Pg93TFY-32KCVpW84>
```

```{block2, type='youtube'}
<https://www.youtube.com/watch?v=IL-2Su9YQ60&index=18&list=PL018X5Hlr4RkgE65Pg93TFY-32KCVpW84>
```

```{block2, type='youtube'}
<https://www.youtube.com/watch?v=HM0oNfNZQlY&index=19&list=PL018X5Hlr4RkgE65Pg93TFY-32KCVpW84>
```


Suppose I am interested in the prevalence of left handedness and wish
to determine whether there is a difference between boys and girls.  I
might survey some students and end up with a dataset like the
following:


```{r}
a <- matrix(c(5,2,6,14),2,2,byrow=TRUE)   # define matrix 
dimnames(a) <- list(gender=c("M","F"),lefthanded=c(T,F))  # display only
a
```

Such a dataset is known as a contingency table and is surprisingly
common in statistics.  I would say that over half of the statistical
analyses I perform for clients is on contingency tables like the one
above.

Is there any evidence that handedness differs by gender?  Just looking
at the table, we can see that the majority of males are left handed
and the majority of females are right handed.  We need to provide
statistical analysis of this question which means providing a p-value.

The starting point would be to provide a null hypothesis, which is
that handedness is independent of gender.  We would thus have
$H_0\colon\prob{L|M}=\prob{L}$.  This is mathematically identical to
the assertion that $\prob{L|M}=\prob{L|F}$.  It certainly looks as
though there is a difference, but to quantify this we need a more
careful analysis.

## The hypergeometric distribution

```{block2, type='wikipedia'}
<https://en.wikipedia.org/wiki/Hypergeometric_distribution>
```

(note that the wikipedia page has different parameterizations from the
R help page.  Here, I will follow the notation used in R, but be aware
that other conventions exist).  Suppose we have an urn with $m$ white
and $n$ black balls, and we draw out $k$ balls and write down their
colour.  Note carefully that we are drawing *without replacement*: we
draw a ball from the urn and then do not put it back in the urn (if we
were to replace the ball in the urn, the number of white balls would
be binomial with size $k$ and probability $\frac{m}{m+n}$).

What is the distribution of the number of white balls drawn without
replacement?  Well, it can be shown that

\begin{equation}
\prob{X=x}=
\frac{{m\choose x}{n \choose k-x}}{{m+n\choose k}}
\end{equation}

where $X$ is the random variable corresponding to the number of white
balls drawn and $x$ is an integer.  The R idiom for this is
```dhyper()``` which calculates the density function.  We have the
standard quartet of functions: ```rhyper()```, ```dhyper()```,
```qhyper()``` and ```phyper()``` which are documented together.  The
```dhyper()``` function takes a number of arguments.

Consult the R help page for details, but the basic command is
```dhyper(x,m,n,k)``` where $x$ is the number of white balls drawn,
$m$ number of white balls in the urn, $n$ number of black balls in the
urn, and $k$ is the number of balls drawn.

We can make the following mapping between drawing balls out of a urn
and left/right male/female dataset as follows:

* white ball $\longrightarrow$ left handed
* black ball $\longrightarrow$ right handed
* drawn from urn $\longrightarrow$ male
* left in urn $\longrightarrow$ female

Alternatively we might say 

* white ball $\longrightarrow$ male
* black ball $\longrightarrow$ female
* drawn from urn $\longrightarrow$ left handed
* left in urn $\longrightarrow$ right handed

Taking the second  of the two identifications, we have

* number of white balls = number of males =5+2=7
* number of black balls = number of females = 6+14=20
* drawn from urn = left handed = 5+6=11
* left in urn = right handed = 2+14=15

With this identification, the *observation* is the number of white
balls drawn from the urn (= the number of left handed males) which is
5.

The probability of this occurring if the null is true is

```{r}
dhyper(5,7,20,11)
```

Recall the definition of p-value: "the probability, if the null is
true, of obtaining the observation or an observation more extreme".
In this case (if we use a one-sided test), "more extreme" means
greater than the observation of 5; and we know that the number of left
handed males (=white balls drawn) cannot be larger than 7.  So the
p-value is

```{r}
sum(dhyper(5:7,7,20,11))
```

Showing that the p-value of about 7\% is not significant (the
professional uses ```phyper(4,7,20,11,lower.tail=FALSE)```).  We can
get a visual impression easily, using R; study the idiom carefully:

```{r}
x <- 0:7  # define x-axis
plot(x,dhyper(x,7,20,11),type='h',lwd=4,col=c(rep("black",5),rep("red",3))) 
legend("topright",lwd=4, col=c("black","red"),legend=c("regular points","more extreme"))
```

Fortunately, R can deal with all this with the ```fisher.test()``` function:

```{r}
fisher.test(a, alternative="greater")
```

(recall that R variable ```a``` was defined at the beginning of the
chapter).