# Quantile methods {#quantile}

```{block2, type='wikipedia'}
<https://en.wikipedia.org/wiki/Quantile_function>
```

```{block2, type='youtube'}
<https://www.youtube.com/watch?v=wJQATecNZHk&index=36&list=PL018X5Hlr4RkgE65Pg93TFY-32KCVpW84>
```

You will recall the deficiencies of the histogram for independent
observations.   Consider, for example, the following diagram:


```{r}
set.seed(0)
d <- rnorm(30)
hist(d)
```

In the above, we see that the precise values of the observations are
hidden by the bins of the histogram.  We can of course plot the entire dataset:

```{r}
plot(d)
```

but this introduces a spurious and possibly misleading aspect to the
plot, namely the order in which the datapoints are plotted.  Because
the observations are independent, the order in which they appear can
should have no effect on any inferences we make.  We can exploit this
fact by plotting the *order statistic*, which is the dataset we
observe but sorted from smallest to largest:


```{r}
plot(sort(d))
abline(v=15.5)
```

In the figure above, the median has been shown as a vertical line.
However, it is more convenient to transpose the axes, and the R idiom
for this is `ecdf()` [the letters stand for empirical cumulative
distribution function]:


```{r}
plot(ecdf(d))
```

In the figure above, the vertical axis has been scaled to represent a
probability rather than a count, and horizontal lines have been added
for convenience.  The function shown is an empirical [data-driven]
approximation to the cumulative distribution function,
$F(x)=\prob{X\leq x}$.  Note that the points appear at the *left* of
the horizontal lines, showing that the function approximated is
$\prob{X\leq x}$ and not $\prob{X<x}$.

We can use R to investigate the properties of the empirical cumulative
distribution function for Gaussian data with a large number of
observations:

```{r}
plot(ecdf(rnorm(1000)))
x <- seq(from=-3,to=3,len=100)
points(x,pnorm(x),col="red",type="l")
legend("topleft",col=c("black","red"),legend=c("empirical","theoretical"))
```

In the figure above, see how the distribution function `pnorm()`
of the Gaussian is approximated.  However, the theoretical (red) line
is not straight, and this makes visual assessment harder.  It would be
a nice if we could distort the horizontal axis so that the theoretical
line is straight, and this would make visual comparison easier.  The
appropriate way to do this is to transform the horizontal axis with
the `pnorm()` function.  The technique works nicely even if the
mean and standard deviation are modified; the appropriate R idiom for
this is `qqnorm()`:

```{r}
qqnorm(rnorm(40,mean=4,sd=2))
abline(4,2)
```

In the figure above, see how the data fall approximately on a straight
line; it turns out that the mean is approximated by the intercept and
the standard deviation by the slope.  The exact values are drawn on
the diagram as the diagonal line.  The `qqnorm()` function may be
used to detect non-normality.  Suppose we sample from a chi-squared
distribution:

```{r}
qqnorm(rchisq(100,df=4))
```

In the figure above, we see a marked curve on the line indicating that
the distribution is not Gaussian.

## Quantile-quantile plots

We may apply similar techniques to compare two datasets without losing
any data as would be the case if we simply drew two histograms.  By
plotting the two datasets' order statistics against each other, (one
on the $x$ axis and one on the $y$ axis), we have a powerful visual tool
to compare two datasets:

```{r}
plot(sort(rnorm(50)),sort(rnorm(50)))
abline(0,1)
```

In the above, we are plotting the smallest observation against the
smallest, the second smallest against the second smallest, and so on,
up to the largest observation.  If the two datasets are drawn from the
same distribution then the points will be close to the $45^\circ$
line, as they are in this case.

If we have different numbers of observations then the R idiom gets trickier and we can use the builtin function `qqplot()`:

```{r}
qqplot(rnorm(20),rnorm(40,mean=1,sd=2))
abline(0,1)
```

In the above figure, the points are not close to the $45^\circ$ line
showing that the two distributions are different (note that the
function deals with different sample sizes: 20 and 40).  However, the
points do fall on a *straight* line which shows that the two
distributions are related by a linear transform, in this case we can
see that the means and standard deviations differ.  If the
distributions are of different shapes, then this shows up as a curved
line:


```{r}
qqplot(rnorm(90),rchisq(99,df=3))
```

In the above, we plot quantiles of normal random data against
chi-squared random data, and because these are different distributions
we see a curved line.  We can use these techniques in a more practical
setting.  Suppose we have two lakes and we sample fish in the lakes,
measuring their lengths:

```{r}
fish_a <-
c(31.2, 26.3, 24.5, 25.9, 29.5, 29.1, 36.7, 28.3, 32.2, 29.0, 25.4, 
  25.2, 37.9, 35.1, 24.7, 34.9, 29.6, 31.8, 26.0, 25.9, 25.5, 35.3, 
  25.7, 28.3, 32.4, 39.2, 33.5, 26.4, 41.3, 37.1, 44.2, 49.0, 46.2, 
  41.1, 55.3, 47.6, 44.7, 45.5, 45.4, 46.7, 47.0, 46.7, 46.4, 43.6, 
  43.0, 47.9, 45.5, 46.0, 42.9, 43.4, 48.1, 38.5, 46.5, 45.4, 46.8, 
  46.8, 50.6, 43.2, 42.9, 45.5, 49.1, 47.7, 49, 46.7, 47.2)
fish_b <-
c(44.7, 42.5, 38.9, 38.9, 28.8, 27.6, 44.8, 56.6, 51.3, 40.2, 35.4,
  39.7, 47.2, 40.1, 36.6, 37.2, 33.5, 42.8, 42.3, 26.5, 37.7, 28.3,
  47.2, 37.2, 41.3, 29.2, 32.4, 28.6, 30.0, 38.1, 41.4, 36.6, 27.0,
  41.1, 50.8, 36.3, 47.5, 47.4, 32.6, 41.9, 32.5, 32.5, 48.3, 37.9,
  47.4, 23.4, 30.6, 23.8, 46.1, 29.9)
qqplot(fish_a, fish_b)
abline(0,1)
```

The above figure gives evidence that the two populations are
different.  To provide quantitative evidence for a difference, we need
techniques discussed in the next chapter.