---
title: "Poisson distribution and Bayes's theorem"
output:
  pdf_document: default
params:
  cutout: TRUE
---

```{r setup, include=TRUE,echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here is assignment 3.  You are free to
discuss the issues with your classmates but copying is not acceptable;
see the academic integrity guidelines.

This assessment has five sections.  All have equal mark value.

Some sections below contain an excellence question.  These are
intended to challenge the very top students.  Have a go at these, but
be aware that they are intended as thought-provoking cues, rather than
course material reproduction.

# Poisson distribution part 1

Consider a random variable with a Poisson distribution with mean
$\lambda = 6.5$.

* Plot the probability mass function for this random variable

```{r, include = !params$cutout}
a <- 0:20
plot(a,dpois(a,6.5),type='h')
```


* Verify that the probabilities sum to 1

```{r, include = !params$cutout}
sum(dpois(0:10000,6.5)) # should equal 1 very closely
```


* Sample from the distribution and plot a table of your results

```{r, include = !params$cutout}
rpois(10,6.5)
plot(table(rpois(1e6,6.5)))
```


* Excellence question: compare this random variable with
  $\operatorname{Bin}(20,0.325)$ (which has the same mean).  NB: for
  high marks you will have to interpret ``compare'' intelligently.

```{r, include = !params$cutout}
par(pty='s')
bin <- dbinom(0:20,20,0.325)
fish <- dpois(0:20,6.5)
plot(bin,fish,pch=16,type="p",asp=1)
abline(0,1)
plot(0:20,bin,type='b',col='black')
points(0:20,fish,type='b',col='red')
legend("topright",col=c("black","red"),legend=c("binomial","poisson"),lty=1,pch=1)
```

# Poisson distribution part 2

A random variable $X$ is known to be drawn from a Poisson
distribution.  An observation is made: $n=6$ is obtained.

* State the precise definition of $p$-value and perform a one-sided
  test for the null that $\lambda = 3.3$.

```{r, include = !params$cutout}
# the probability, if the null is true, of obtaining the observation or
# an observation more extreme


1-sum(dpois(0:5,3.3))
ppois(5,3.3,lower.tail=FALSE)
# pvalue exceeds 0.05, no strong evidence
```



* Calculate the likelihood ratio for $H_1\colon\lambda = 4.1$ and
  $H_2\colon\lambda=2.2$.
  
```{r, include = !params$cutout}
dpois(6,4.1)/dpois(6,2.2)
```
  
* Plot a log-likelihood function in the range $2\leq\lambda\leq 14$.

```{r, include = !params$cutout}
l <- seq(from=2,to=14,len=100)
plot(l,dpois(6,l,log=TRUE))
```

* Calculate a credible interval for $\lambda$, using a
  2-units-of-support criterion [hint: follow page 67 of the course manual].
  An approximate answer is fine
  
```{r,include = !params$cutout}
L <- dpois(6,l,log=TRUE) - dpois(6,6,TRUE)
plot(l,L)
abline(h=c(0,-2))
## from the graph, about 2.3-12
```  
* Excellence question: calculate a confidence interval as per the end
  of chapter 3 in the course manual; compare the confidence interval
  with the credible interval above.  One decimal place sufficient.

```{r, include = !params$cutout}
1-sum(dpois(0:5,2.6))
1-sum(dpois(0:5,2.7))

sum(dpois(0:5,10.5))
sum(dpois(0:5,10.6))
## CI of about 2.6-10.6, narrower than credible interval
## 

## more accurate:
f <- function(x){ppois(5,x,lower.tail=F)-0.05}
uniroot(f,2:3)$root
g <- function(x){ppois(5,x,lower.tail=T)-0.05}
uniroot(g,10:11)$root
```

# Poisson distribution part 3

Give an example drawn from your daily life of a random variable drawn
from a Poisson distribution.  You will recall examples from the manual
and lectures: but do not use these examples!  USe these examples to
guide your understanding but I am looking for your own original
examples.

* State _clearly_ in what way it is a limiting form of a binomial
  distribution with large $n$ and small $p$.
* Give rough estimates for $n$ and $p$ [an order-of-magnitude estimate
  is fine]
* Give a reasonable value for $\lambda$ (a rough value is fine)
* Using your value for $\lambda$, sample from your Poisson
  distribution using `rpois(10,lambda)`.  Interpret the meaning of the
  10 random numbers.  What does the first number mean?  What does the
  second number mean?
* Excellence question: discuss whether your distribution really is a
  Poisson distribution.  Are the statistical assumptions justified?
  Discuss.


# Bayesian analysis part 1

A certain person buys a houseplant.  She buys either a cactus or an
orchid.  From long experience I can say that it is a cactus with
probability 0.4, or an orchid with probability 0.6.  She then goes on
holiday and does not water her plants.  The plants get too dry and
might die.  Under these circumstances, a cactus will die with
probability 0.05 and an orchid with probability 0.24.

The plant dies.  Apply Bayes's theorem and show your working.  What is
the posterior probability that it was a cactus?

Excellence question: consider the probability of the plant being a
cactus.  Discuss how this probability changes from prior to posterior
in the light of your observation and whether this is consistent with
your intuition.


```{r, include = !params$cutout}
cactus <- 0.4*0.05
orchid <- 0.6*0.24
cactus/(cactus  + orchid)
#  probability decreases from 40% to 12% as cacti do not die but orchids do
```

# Bayesian analysis part 2

Give a situation from your everyday life and apply Bayes's theorem to
it.  Use two or three hypotheses.  State your hypotheses _clearly_.
State your prior probabilities _clearly_.  State your data _clearly_.
State your likelihood function _clearly_.  Use Bayes's theorem to
calculate posterior probabilities for your hypotheses and state your
posterior probabilities _clearly_.  For an excellence grade, interpret
the difference between the prior and posterior probabilities using
common-sense or intuitive reasoning.

Credit will be given for the use of a plausible and novel scenario,
the inventiveness shown, and the appropriateness of Bayes's theorem.


