---
title: "The binomial distribution"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The binomial distribution

This assessment requires you to use R to describe some properties of
the binomial distribution.  Each part has equal marks value; attempt
them all.  Submit via Canvas.

Before you start, execute the following R code which will give you a
size and a probability of success.  Students who use my ID will
receive zero credit.


```{r}
set.seed(1266402) # This is my student ID: replace with yours!
sample(20:30,1)   # use this for the size
rbeta(1,2,2)      # use this for the probability 'p' of success
```

Above we see that (for me) the size of the distribution is $n=24$ and
the probability of success is $p=0.3974$ (use 4 decimal places).
Remember to replace my student ID with yours to get your personal
value for size and probability.  For the questions below, use the
course manual as a guide.

* Plot the probability mass function for your distribution
* Check that the probabilities sum to one
* Use `rbinom()` to generate a random sample from your distribution
  and give a frequency plot of your observations
* Define your own function [call it `mydbinom()`] that calculates
  `dbinom()` and compare with R's `dbinom()`.  If you can, find a
  value for `n,r,p` for which `mydinom(n,r,p)` differs from
  `dbinom(n,r,p)`.
* For extra credit, sample from the appropriate _Bernoulli_
  distribution using R, and use this to verify that `dbinom()` is
  approximately correct.




