---
title: "Hypothesis testing"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This assessment requires you to use R to perform some hypothesis
testing and work with some properties of the Gaussian distribution.
Each question is of equal mark value.

First, execute the following code which will give you a mean and
standard deviation.  Students who use my ID will receive zero credit.

```{r}
set.seed(1266402)     # This is my student ID: replace with yours!
rnorm(1,10,2)         # Use this for the mean
sqrt(rchisq(1,df=2))  # Use this for the standard deviation
```

# Gaussian distribution

Above we see that the mean of the distribution is $\mu=11.40$ and the
standard deviation is about 0.704.  But use your values, not mine.
For the questions below, use the course manual as a guide.

* Plot the probability density function for your distribution

* If random variable $X$ is drawn from your Gaussian distribution,
  find the probability that $X$ lies in the range
  $(\mu-\sigma,\mu+2\sigma)$
* Use `rnorm()` to generate a random sample from your distribution
  and give  a histogram  of your observations
* For bonus credit, find the steepest point of the standard normal
  distribution (this is a numerical exercise, not calculus).


#  Student t-test

The average weight of a banana is 100 grams.  An agricultural
scientist buys bananas from a supermarket.  Their weight, in grams, is
as follows:
  
```{r}
w <- c(102.5, 102.9, 99.9, 101.9, 101.3, 98.9, 96.5, 97.7, 101.8, 
100.3)
```

She suspects that this sample of bananas is heavier than average and
wonders if this supermarket is selling bananas that are heavier than
the notional 100g mean weight.

 *  State a sensible null hypothesis
 *  State the precise definition of $p$-value and explain what
    ``more extreme'' means in this context
 *  Is a one-sided or two-sided test needed? justify
 *  Perform a student $t$-test using R and interpret
 *  Give a 95\% and 99\% confidence interval for the mean.
 *  For bonus credit, perform a $Z$ test and account for any differences you find


# Hypothesis testing: binomial distribution

A NZ scientist is studying a particular type of plant that
always has exactly 7 flowers on it.  Each flower may be
fertilized or unfertilized.  The probability of a flower being
fertilized is known to be $0.4$; flowers are independent of one
another and you may use the binomial distribution.

  * Using `dbinom()` or otherwise, what is the probability
    that exactly one flower on a particular plant is fertilized?
  * What is the probability that three or more flowers are fertilized?
  * A plant is found to be very near a beehive, and it is
    suspected that bees fertilize this kind of plant.  The scientist
    finds that $5$ flowers are fertilized.
  * State a sensible null hypothesis
  * state the precise definition of $p$-value and define what "more extreme" means in this context.  
  * One-sided or two sided?  Justify
  * Calculate the $p$-value.  Is it significant? 
  * Interpret your finding in a way that a busy entomologist, who is not a statistician, could understand


