---
title: "Hypothesis testing"
output:
  pdf_document: default
params:
  cutout: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This assessment requires you to use R to perform some hypothesis
testing and work with some properties of the Gaussian distribution.
Each question is of equal mark value.

First, execute the following code which will give you a mean and
standard deviation.  Students who use my ID will receive zero credit.

```{r}
set.seed(1266402)     # This is my student ID: replace with yours!
rnorm(1,10,2)         # Use this for the mean
sqrt(rchisq(1,df=2))  # Use this for the standard deviation
```

# Gaussian distribution

Above we see that the mean of the distribution is $\mu=11.40$ and the
standard deviation is about 0.704.  But use your values, not mine.
For the questions below, use the course manual as a guide.

* Plot the probability density function for your distribution

```{r, include = !params$cutout}
mu <- 10.4
sigma <- 0.704
x <- seq(from=8,to=13,len=100)
plot(x,dnorm(x,mean=mu,sd=sigma))
```

* If random variable $X$ is drawn from your Gaussian distribution,
  find the probability that $X$ lies in the range


  $(\mu-\sigma,\mu+2\sigma)$

```{r, include = !params$cutout}
pnorm(mu+2*sigma,mu,sigma) - pnorm(mu-2*sigma,mu,sigma)
pnorm(2)-pnorm(-2)   # slick
x <- rnorm(1e6,mu,sigma)

# numerical verification (not needed for full marks):
sum((x<mu + 2*sigma) &(x>mu -2*sigma))/1e6 
```

* Use `rnorm()` to generate a random sample from your distribution
  and give  a histogram  of your observations

```{r, include = !params$cutout}
hist(rnorm(1e6,mu,sigma))
```

* For bonus credit, find the steepest point of the standard normal
  distribution (this is a numerical exercise, not calculus).

```{r, include = !params$cutout}
x <- seq(from=-2,to=2,len=1000)
grad <- diff(dnorm(x))
steepest_increasing_point <- x[which.max(c(0,grad))]
steepest_decreasing_point <- x[which.min(c(0,grad))]
steepest_increasing_point
steepest_decreasing_point

plot(x,dnorm(x))
abline(v=c(steepest_increasing_point,steepest_decreasing_point))
```

#  Student t-test

The average weight of a banana is 100 grams.  An agricultural
scientist buys bananas from a supermarket.  Their weight, in grams, is
as follows:
  
```{r}
w <- c(102.5, 102.9, 99.9, 101.9, 101.3, 98.9, 96.5, 97.7, 101.8, 
100.3)
```

She suspects that this sample of bananas is heavier than average and
wonders if this supermarket is selling bananas that are heavier than
the notional 100g mean weight.

 *  State a sensible null hypothesis

```{r, include = !params$cutout}
# Null: mean weight of this sample is 100g
```

 *  State the precise definition of $p$-value and explain what
    ``more extreme'' means in this context

```{r, include = !params$cutout}
# p-value equals the probability, if the null is true, of
# obtaining the observation or an observation more extreme.
# Here, the observation is the sample mean and "more extreme"
# means "greater than the observed sample mean"
```

 *  Is a one-sided or two-sided test needed? justify

```{r, include = !params$cutout}
# One-sided as we suspect the bananas are heavier
```

 *  Perform a student $t$-test using R and interpret

```{r, include = !params$cutout}
t.test(w,mu=100,alternative="greater")
```

 *  Give a 95\% and 99\% confidence interval for the mean.

```{r, include = !params$cutout}
t.test(w,mu=100,alternative="greater",conf.level=0.95)$conf.int
t.test(w,mu=100,alternative="greater",conf.level=0.99)$conf.int
```

 *  For bonus credit, perform a $Z$ test and account for any differences you find

```{r, include = !params$cutout}
Z <- (mean(w)-100)/sd(w)
pnorm(Z,lower.tail=FALSE)

# pvalue higher than student test, consistent with heavier tails of
# the student t distribution [give credit for any remotely plausible
# reasoning here ... note that I do not 100% buy the rationale for the
# Student t test when compared with the Z test]
```


# Hypothesis testing: binomial distribution

A NZ scientist is studying a particular type of plant that
always has exactly 7 flowers on it.  Each flower may be
fertilized or unfertilized.  The probability of a flower being
fertilized is known to be $0.4$; flowers are independent of one
another and you may use the binomial distribution.

  * Using `dbinom()` or otherwise, what is the probability
    that exactly one flower on a particular plant is fertilized?

```{r, include = !params$cutout}
dbinom(1,7,0.4)
```

  * What is the probability that three or more flowers are fertilized?

```{r, include = !params$cutout}
sum(dbinom(3:7,7,0.4))
pbinom(2,7,0.4,lower.tail=FALSE)
# numerical verification (not needed for full credit but good practice):
table(rbinom(1e6,7,0.4) >=3)
```

  * A plant is found to be very near a beehive, and it is
    suspected that bees fertilize this kind of plant.  The scientist
    finds that $5$ flowers are fertilized.
  * State a sensible null hypothesis

```{r, include = !params$cutout}
# Null: the beehive has no effect and p=0.4
```

  * state the precise definition of $p$-value and define what "more extreme" means in this context.

```{r, include = !params$cutout}
# p-value equals the probability, if the null is true, of
# obtaining the observation or an observation more extreme.
# Here, the observation is the sample mean and "more extreme"
# means "greater than the observed sample mean"
```

  * One-sided or two sided?  Justify

```{r, include = !params$cutout}

# One-sided, bees increase p
```

  * Calculate the $p$-value.  Is it significant? 

```{r, include = !params$cutout}
sum(dbinom(5:7,7,0.4))
# pvalue exceeds 0.05, not significant
```


  * Interpret your finding in a way that a busy entomologist, who is not a statistician, could understand

```{r, include = !params$cutout}
# No evidence that the beehive affects pollination probabilities
```


