---
title: "The Gaussian distribution"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Gaussian or normal distribution

This assessment requires you to use R to describe some properties of
the Gaussian distribution.

Execute the following code which will give you a mean and standard
deviation.  Students who use my ID will receive zero credit.


```{r}
set.seed(1266402)     # This is my student ID: replace with yours!
rnorm(1,10,2)         # Use this for the mean
sqrt(rchisq(1,df=2))  # Use this for the standard deviation
```

Above we see that the mean of the distribution is $\mu=11.40$ and the
probability of success is $p=0.5656$ (use 4 sig figs).  For the
questions below, use the course manual as a guide.

* Plot the probability density function for your distribution

* If random variable $X$ is drawn from your Gaussian distribution,
  find the probability that $X$ lies in the range
  $(\mu-\sigma,\mu+2\sigma)$
* Use `rnorm()` to generate a random sample from your distribution
  and give  a histogram  of your observations
* Verify that `dnorm()` matches the density function given by

$f(x) = \frac{\exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right)}{\sigma\sqrt{2\pi}}$

* For extra credit, find the steepest point of the standard normal distribution
