---
title: "Final assessment"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(include = FALSE)  # change to TRUE when rendering solutions; this changes
                                       # the default visibility (which is overridden by explicitly
                                       # setting  include=TRUE in chunk options) 

```

# Final assessment:  stat601/stat805/comp616

This assessment requires you to use R to demonstrate your
understanding of probability and statistics.  Answer each question and
show your working, including your R code.  There are four questions,
each of 25 marks.


## Question 1


 The temperature of a certain city during the year can be modelled reasonable well
 with a Normal distribution with mean 25 degrees Celsius and standard deviation 4
 degrees Celsius.

* What is the probability that in a random day the temperature is at most 21C?
* Verify your calculation via simulation.

```{r}
mu <- 25
sd <- 4
pnorm(21, mu, sd)
mean(rnorm(1e5, mu, sd)<= 21)
```

* What is the probability that in a random day the temperature is at least 28C?
* Verify your calculation via simulation.

```{r}
1-pnorm(28, mu, sd);
mean(rnorm(1e5, mu, sd) >= 28);
```

* What is the probability that in a random day the probability is between 28C and 32C?
* Verify your calculation via simulation.

```{r}
pnorm(32, mu, sd) - pnorm(28, mu, sd);
x = rnorm(1e5, mu, sd);
mean( x >= 28 & x<= 32)
```

[excellence] We choose 14 days randomly (assume independence between the days).

* What is the probability that the temperature is greater than 30C in at least five days, chosen randomly from the 14?
* Verify the probability through simulation.

```{r}
n = 14;
p = 1 - pnorm(30, mu, sd);
sum(dbinom(5:14, n, p))
mean(rbinom(1e5, n, p)>=5);
```

* How many days with a temperature greater than 30C would you expect?    Verify your answer via simulation.


```{r}
n * p;
mean(rbinom(1e5, n, p));
```


## Question 2

A vetinarian is investigating a particular disease of cows.  She is
interested in whether English cows differ from Scottish cows in their
probability of suffering from this particular disease.  A sample of 15
cows is taken, of which 9 are English and 6 are Scottish.  It is not
known if English cows are more or less likely to have this disease.
In the sample, there are 10 healthy cows, of which 8
are English.  We may represent the data in R as follows:
  

```{r,include=TRUE}
x <- 8  # healthy English (white drawn)
k <- 9  # English (drawn)
n <- 5  # diseased (black)
m <- 10 # healthy (white)

a <- matrix(c(x,k-x,m-x,n-k+x),2,2)
dimnames(a) <- list(
    country=c("England","Scotland"),state=c("healthy","diseased")
)
a
```

  * State a sensible null hypothesis
  * State the precise definition of $p$-value and explain what
    ``more extreme'' means in this context
  * Is a one-sided or two-sided test needed? justify
  * Perform a Fisher test using `fisher.test()` and interpret
  * Estimate the probability of: (a), an English, and (b), a
    Scottish cow, having the disease
  * give a 95\% Gaussian confidence interval for the probabilities of English and Scottish cows having the disease (use the normal approximation)
  * excellence: verify your $p$-value given by `fisher.test()` using `dhyper()` directly.     


```{r}
# Null :probability of an English cow having the disease is equal to the
# probability of a Scottish cow having the disease
# p-value: probability, if the null is true, of obtaining the observation
# or an observation more extreme
# Here the observation is the number of healthy English cows and "more 
# extreme" means "a lower probability under the null".
fisher.test(a)
f <- function(a,b){qnorm(c(0.025,0.5,0.975),a/(a+b),sqrt((a*b)/(a+b)^3))}

f(x,m-x)      # English
c(healthy=k-x,diseased=n-k+x)
f(k-x,n-k+x)  # Scottish

# excellence:
fishp <- function(x){dhyper(x,m,n,9)}
p <- fishp(1:10)
sum(p[p<=fishp(8)])
```

## Question 3

A computer scientist in Auckland has a large number of computers in
his laboratory and suspects that more failures occur when the
temperature outside is higher.

He collects data on 23 days:


```{r,include=TRUE}
temp <- c(17.1, 19.2, 15.2, 18.1, 15, 17.8, 17.8, 15.2, 17.4, 15.7, 17.8, 
16.9, 17.1, 17, 18, 15.9, 16.5, 17.3, 16.7, 15.9, 16.1, 19, 16)
fail <- c(28, 30, 27, 33, 27, 34, 31, 29, 29, 26, 33, 31, 27, 31, 32, 
28, 30, 33, 32, 26, 30, 34, 27)
```

Above, `temp` is the temperature in centigrade, and `fail` the number
of failures on each day.

  * State a sensible null hypothesis
  * State the precise definition of $p$-value and explain what
    ``more extreme'' means in this context
  * Is a one-sided or two-sided test needed? justify
  * Perform a linear regression using R and interpret.  Is there
    evidence that more computer failures occur during hotter weather?


```{r}
# Null: no relation between failures and temperature
# pvalue is the probability, if the null is true, of obtaining the observation
# or an observation more extreme. Here the observation is the fitted slope and
# "more extreme" means "higher fitted slope"
plot(fail~temp,xlab="temperature (C)",ylab="failures")
fit <- lm(fail~temp)
abline(fit)
summary(fit)  # one-sided test [slope of curve >0]; halve p-value
```


## Question 4

Around 37 degrees Centigrade is normal body temperature for children
and adults.  _Fever_ is when your core body temperature is higher than
38 degrees centigrade.  A certain medicine that aims to reduce body
temperature has been applied to 80 patients that had a fever.  The
temperatures after the medicine was administrated are the following:


```{r,include=TRUE}
y  <-  scan( text = "
37.3 37.5 37.2 38.0 37.6 37.2 37.6 37.7 37.7 37.4 37.9 37.6 37.3
36.8 37.8 37.5 37.5 37.8 37.7 37.7 37.8 37.7 37.5 36.9 37.7 37.5
37.4 37.0 37.3 37.6 37.9 37.5 37.6 37.5 37.1 37.4 37.4 37.5 37.8
37.7 37.4 37.4 37.7 37.7 37.3 37.3 37.6 37.7 37.5 37.8 37.6 37.3
37.6 37.2 37.9 38.1 37.4 37.2 37.7 37.4 38.2 37.5 37.7 37.5 37.3
37.5 36.9 37.9 37.5 38.1 37.6 37.3 37.7 37.2 37.1 37.6 37.4 37.5
37.5 37.3")
```

We consider that the medicine is _effective_ if the temperature is
lower or equal to 38.  Perform a Bayesian analysis assuming a
skeptical position: we seek the probability that the medicine is
effective.

* Define the random variable and its distribution.
* Define the prior and justify your choice.
* Define the posterior distribution.
* Plot the prior and posterior in a graph.  Comment on it.
* Calculate a 95% credible interval and interpret it.

```{r}
# we seek p, the probability that T<38
table(y<38)
## uniform prior, alpha=beta=1:
p <- seq(from=0,to=1,len=100)
alpha <- 1
beta <- 1
plot(p,dbeta(p,alpha,beta),type='l',col='black',ylim=c(0,20))
points(p,dbeta(p,alpha + 76 ,beta+4),type='l',col='red')
legend("topleft",lty=1,col=c("black","red"),legend=c("prior","posterior"))
qbeta(c(0.025,0.975),alpha+76,beta+4)
```


