---
title: 'Assignment (statistics): Fisher''s exact test, Pearson''s chi square test, linear regression and logistic regression'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE) # change to TRUE when rendering solutions; this changes
                                       # the default visibility (which is overridden by explicitly
                                       # setting  include=FALSE in chunk options) 
```

# Instructions

Answer the following four questions, which have equal mark value.
Show your working by including output from your R session.



# Question 1 part (a)

Consider the following dataset drawn from AUT student services:

```{r}
M <- matrix(c(10,2,11,7),2,2)
dimnames(M) <- list(OS=c("windows","mac"),major=c("science","arts"))
M
```

we suspect arts students are more likely to use a mac than science students.

* State your null clearly
* State the precise definition of p-value
* state what "more extreme" means here
* use `fisher.test()`, calculate your pvalue and interpret


```{r}
fisher.test(M,alternative="greater")
# no significant difference
```

# Question 1 part (b)

A medical researcher is investigating whether vitamin C helps to cure
the common cold.  He takes a sample of 20 subjects, of whom
7 are given vitamin C and 13 a placebo (a "placebo" is
a dummy pill that has no medical effect).  The researcher interviews
the subjects after one week, and 9 subjects report that their
cold is cured.  Of these 9 subjects who are cured, 5
were given real vitamin C. 

* State a sensible null hypothesis
* State the precise definition of $p$-value and explain what
    ``more extreme'' means in this context
* Is a one-sided or two-sided test needed? justify
* Estimate the probability of: (a), a person who was given
    vitamin C, and (b), a person given the placebo, of being cured.
* Perform a Fisher test using `fisher.test()` and interpret

```{r}
# p-value is the probability, if the null is true, of obtaining the observation or an observation more extreme.
# Null: vitamin C has no effect on colds.  "more extreme" = more than 5 vit C patients cured
M <- matrix(c(5,4,2,9),2,2)
dimnames(M) <- list(
    given=c("vitamin C","placebo"),state=c("cured","cold")
)
M

fisher.test(M,alternative="greater")  # pvalue < 0.05 -> significant evidence for cure
5/7  # prob of vit C group being cured
4/13 # prob of placebo group being cured
```



# Question 1 part (c)

Give an example of Fisher's exact test in your daily life.  Give a 2x2
contingency table, with labelled rows and columns.  State your null
_clearly_, and your alternative.  State and justify your use of a
one-sided or two-sided test.  Carry out your test, report the p-value,
and interpret.  Excellence question: find the "most extreme"
observation that is consistent with your marginal totals.



# Question 2 part (a)

62 students in a certain statistics class are asked which is their
favourite colour among Red, Orange, Blue, and Green.  The results are
as follows:

```
o <- c(red = 10, orange = 16, yellow = 24, green = 12)
```

* State a sensible null hypothesis
* State the observed values and the expected values under your null.
* Calculate the Pearson's chi-square statistic `sum((o-e)^2/e)` and show your working; state what its approximate null distribution is
* State the precise definition of p-value and explain what "more extreme" means in this context.
* Calculate the pvalue using `pchisq()` and your result from above, and interpret.
* Excellence question: using random simulation, or otherwise, estimate
  the probability that all four colours have different numbers of
  students.


```{r}
o <- c(red = 10, orange = 16, yellow = 24, green = 12)  # observed
e <- mean(o)                                            # expected
s <- sum((o-e)^2/e)                                     # Pearson's chi-square statistic
## Null: equal probability of choosing each colour
## p-value is the probability, if the null is true, of obtaining the observation
## or an observation more extreme.
## Here "more extreme" means "a larger chi-square statistic"
pchisq(s,df=3,lower.tail=F)                             # pvalue
 ## not significant (exceeds 0.05)
```


```{r}
table(replicate(1000,max(table(table(sample(1:4,62,replace=T))))))
# About 64%
```



# Question 2 part (b)


Give an example of Pearson's chi-square test from your daily life.
Have at least three categories.  State your observation _clearly_,
state your null and expected values _clearly_.  Calculate and state
the value of the chi-square statistic $B$.  State and justify the null
distribution of $B$ (including number of degrees of freedom), give a
$p$-value.  Interpret your findings.  Use clear, succinct, language.


# Question 3

[remember the golden rule: without a scattergraph, there is no point
in performing any regression]

A computer scientist in Auckland has a large number of computers in
his laboratory and suspects that more failures occur when the
temperature outside is higher.

The dataset is as follows:

`temp <- c(16.1, 18.2, 16.5, 14.1, 15.2, 17.2, 17.9, 17.5, 16.8, 14.3)`

`fail <- c(9.90, 9.21, 10.01, 8.22, 9.33, 8.89, 9.58, 10.97, 10.38, 8.96)`

The first dataset is the temperature in Auckland in Centigrade, the
second is the percentage of computers that have a failure.

* Remembering the golden rule, perform a linear regression of the above dataset.
* State your null clearly.
* One-sided or two-sided test?   Justify
* Interpret your result using non-statistical language.


```{r}
temp <- c(16.1, 18.2, 16.5, 14.1, 15.2, 17.2, 17.9, 17.5, 16.8, 14.3)
fail <- c(9.90, 9.21, 10.01, 8.22, 9.33, 8.89, 9.58, 10.97, 10.38, 8.96)
plot(fail~temp)
(fit <- summary(lm(fail~temp)))
abline(fit)
## Null: no relation between computer failure and temperature.
## pvalue = 0.132/2=0.06 [one sided, expect temp to increas failure rate]; not significant.
## No evidence for any association.
```




# Question 4

A researcher is investigating the link between myocardial infarction
(heart attack) and blood pressure.  He takes 30 patients drawn from a
particular demographic and measures their systolic blood pressure.
The results are as follows:

`p <-  c(104, 109, 111, 114, 116, 120, 129, 132, 136, 138,`
`140, 141, 142, 145, 148, 151, 153, 155, 156, 157, 160, `
`161, 162, 166, 170, 173, 177, 193, 194, 195)`

`H <- c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, `
`0, 0, 0, 1, 0, 0, 0, 1, 0, 1, 1, 0, 1, 1, 1)`


Above, `p` is the systolic blood pressure in mmHg and `H` shows
whether or not that patient has had a heart attack before.  Use
logistic regression to analyse this dataset.  Can we say that high
blood pressure {\em causes} heart attacks?


```{r}
p <-  c(104, 109, 111, 114, 116, 120, 129, 132, 136, 138, 
140, 141, 142, 145, 148, 151, 153, 155, 156, 157, 160, 
161, 162, 166, 170, 173, 177, 193, 194, 195)

H <- c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 
0, 0, 0, 1, 0, 0, 0, 1, 0, 1, 1, 0, 1, 1, 1)

plot(H~p)
summary(glm(H~p,family="binomial"))

LO <- function(p){log(p/(1-p))}
p <- function(LO){1/(1+exp(-LO))}

x <- 10:200
points(x,p(-15.9 + 0.095*x),type="l")
# pvalue = 0.0127/2 = 0.007 [one-sided, expect high P to increase prob], highly significant.
# High blood pressure is associated with increased risk of heart attack.
# Correlation does not imply causation.
```




