---
title: 'Assignment (statistics): Fisher''s exact test and Pearson''s chi square test'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE) # change to TRUE when rendering solutions; this changes
                                       # the default visibility (which is overridden by explicitly
                                       # setting  include=TRUE in chunk options) 
```

# Instructions

Answer the following questions, which all have equal mark value.  Show
your working by including output from your R session.


# Question 1

Consider the following dataset drawn from AUT student services:

```{r}
M <- matrix(c(10,2,11,7),2,2)
dimnames(M) <- list(OS=c("windows","mac"),major=c("science","arts"))
M
```

we suspect arts students are more likely to use a mac than science students.

* State your null clearly
* State the precise definition of p-value
* state what "more extreme" means here
* use `fisher.test()`, calculate your pvalue and interpret


```{r,include=TRUE}
fisher.test(M,alternative="greater")
# no significant difference
```

# Question 2

A medical researcher is investigating whether vitamin C helps to cure
the common cold.  He takes a sample of 20 subjects, of whom
7 are given vitamin C and 13 a placebo (a "placebo" is
a dummy pill that has no medical effect).  The researcher interviews
the subjects after one week, and 9 subjects report that their
cold is cured.  Of these 9 subjects who are cured, 5
were given real vitamin C.  We may represent the data in R as follows:
  
```{r}
M <- matrix(c(5,4,2,9),2,2)
dimnames(M) <- list(
    given=c("vitamin C","placebo"),state=c("cured","cold")
)
M
```

* State a sensible null hypothesis
* State the precise definition of $p$-value and explain what
    ``more extreme'' means in this context
* Is a one-sided or two-sided test needed? justify
* Estimate the probability of: (a), a person who was given
    vitamin C, and (b), a person given the placebo, of being cured.
* Perform a Fisher test using `fisher.test()` and interpret

```{r,include=TRUE}
# p-value is the probability, if the null is true, of obtaining the observation or an observation more extreme.
# Null: vitamin C has no effect on colds.  "more extreme" = more than 5 vit C patients cured
fisher.test(M,alternative="greater")  # pvalue < 0.05 -> significant evidence for cure
5/7  # prob of vit C group being cured
4/13 # prob of placebo group being cured
```

# Question 3

Consider a situation with 3 white and 5 black balls in a bag.  Four balls are drawn from the bag, without replacement.  Write down every possible sample and calculate its probability.

```{r,include=TRUE}
f <- function(n){
  out <- matrix(c(n,3-n,4-n,1+n),2,2)
  dimnames(out) <- list(cstate=c("drawn","left"),colour=c("white","black"))
  out
}

f(0)
f(1)
f(2)
f(3)
dhyper(0:3,3,5,4)
```

# Question 4

Three AUT students and four UoA students are given a problem in statistics.  All four of the AUT students answer the problem correctly, and none of the UoA students answer correctly.  Discuss.

```{r}
fisher.test(diag(3:4)) # two sided? . . . . . 
# strong evidence
```


# Question 5

Give an example of Fisher's exact test in your daily life.   Give a 2x2 contingency table, with labelled rows and columns.  State your null _clearly_, and your alternative.  State and justify your use of a one-sided or two-sided test.  Carry out your test, report the p-value, and interpret.  Excellence question: find the "most extreme" observation that is consistent with your marginal totals.






