---
title: "Assignment (statistics)"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE) # change to TRUE when rendering solutions; this changes
                                       # the default visibility (which is overridden by explicitly
                                       # setting  include=TRUE in chunk options) 
```

# Instructions

Answer the following FIVE questions, which all have equal mark value.  Show
your working by including output from your R session.


# Question 1

Consider the binomial distribution $X\sim\operatorname{Bin}(n,p)$.

* If $n=9,p=0.3$, what is the probability that $X\leq 4$?  Verify your answer using direct numerical simulation.

```{r}
sum(dbinom(0:4,9,0.3))
pbinom(4,9,0.3)
table(rbinom(1e6,9,0.3) <= 4)/1e6
```

* If $n=100, p=0.4$, what is the probability that $X\geq 50$?  Calculate the exact probability, and the value obtained by 
using the Gaussian approximation to the binomial. 


```{r}
sum(dbinom(50:100,100,0.4))
pnorm(50,100*0.4,sqrt(100*0.4*(1-0.4)),lower.tail=FALSE)
# results quite close.
```

* Consider $X\sim\operatorname{Bin}(9,p)$ where $p$ is unknown.  Plot
  a likelihood function for $p$ for the observation $X=3$.  Indicate
  the maximum likelihood estimate $\hat{p}$ on the graph.
* Calculate
  the likelihood ratio for $H_1\colon p=1/2$ against $H_2\colon
  p=\hat{p}$.

```{r}
p <- seq(from=0,to=1,len=100)
plot(p,dbinom(3,9,p))
abline(v=1/3)
dbinom(3,9,1/3)/dbinom(3,9,1/2)
```

# Question 2

Consider a fire alarm which is intended to alert occupants of a
building to the fact that there is a fire in the vicinity.  The fire
alarm malfunctions occasionally.  Give a brief discussion of the
operation of this fire alarm with regard to type I and type II errors.
State clearly any null hypothesis you use.

```{asis}
null: no fire

alternative: fire

type I: reject null when true, alarm rings when no fire; false alarm

type II: fail to reject null when false, alarm stays silent when there is a fire,
alarm fails to alert occupants.
```

Now suppose that the following probabilities are correct for a given day:

* Prob(fire) =0.0001

* Prob(alarm rings if there is a fire) = 0.9

*  Prob(alarm rings if there is no fire) = 0.1

One day, the alarm rings.  Give a likelihood ratio for your null and
alternative hypotheses.

```{r}
0.9/0.1
```

# Question 3 

Consider the following dataset which refers to the weights, to the nearest kg,
of six people who join a fitness program.  The figures refer to the participants'
weights before and after the program. 

```{r,include=TRUE}
weights_before <- c(bill = 100, baz=67, bob=77, sandy=91, fred=94, kevin=60)
weights_after <- c(bill = 101, baz=62, bob=75, sandy=90, fred=89, kevin=61)
```


* State a sensible null hypothesis
* State the precise definition of $p$-value
* State and justify your use of one- or two-sided tests
* Perform the appropriate t-test and interpret the result
* (harder) Using maximum likelihood estimates for the variance,
  perform a $Z$-test on the same data and compare the result with the
  $t$-test above.


```{r}
t.test(weights_before,weights_after,paired=TRUE,alternative="greater")
```

```{asis}
$p$-value exceeds 0.05, no evidence.

For the rider, there are different acceptable solutions.  Here is one:
```

```{r}
residuals <- weights_before - weights_after
S <- sd(residuals)*sqrt(6/5)
pnorm(mean(residuals),sd=S,lower.tail=FALSE)
```


# Question 4

A sports statistician is studying the number of goals 
scored by the home team in a particular football league.
The results are as follows:

```{r, include=TRUE}
d <- c(goals0=6,goals1=3, goals2=7, goals3=6, goals4=7, goals5=3, goals6=3) 
```

Thus we observe 2 matches with zero goals ("`goals0=2"), 3 matches with one goal
("`goals1=3`"), 7 matches with 3 goals, and so on.

The statistician suspects that the number of goals scored follows a Poisson distribution.

* Give a plausible reason why a Poisson distribution might be appropriate.
* Assess this claim using Pearson's chi-square test.  State your null
  clearly and state the precise definition of p-value.
* Interpret your result.



```{r}
goals <- sum(d *( 0:6))
matches <- sum(d)
(lhat <- goals/matches)
probs <- dpois(0:5,lambda=lhat)
probs <- c(probs,1-sum(probs)) # >= 6
probs
o <- d
e <- probs*matches
o
e
B <- sum((o-e)^2/e)
pchisq(B,df=7-1-1,lower.tail=FALSE)
```


```{asis}
pvalue exceeds 0.05, not significant.
```
  
# Question 5

Two police officers, A and B, conduct roadside sobriety testing of
randomly selected drivers.  Officer A tests 21 drivers of whom 3 fail
the test.  Officer B tests 33 drivers of whom 8 fail.  Is there any
evidence that the police officers differ in their probability of
failing a driver?

* State a sensible null hypothesis
* State the precise definition of $p$-value
* Assess your hypothesis using Fisher's exact test.  Indicate and justify your use
of one-sided or two-sided tests.
* Interpret your result.



```{asis}
    Two-sided; the situation is symmetrical 
```


```{r}
a <- matrix(c(18,25,3,8),2,2)
dimnames(a) <- list(officer=c("A", "B"), test=c("sober","fail"))
a
fisher.test(a,alternative="two.sided")
```

```{asis}
Significant, just.
```

