---
title: 'Assignment: Pearson''s chi-square test'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)  # change to TRUE when rendering solutions; this changes
                                       # the default visibility (which is overridden by explicitly
                                       # setting  include=TRUE in chunk options) 
```

# Instructions

Answer the following questions.  Show your working by including output
from your R session.  All four questions are of equal mark value.

# Question 1.

62 students in a certain statistics class are asked which is their
favourite colour among Red, Orange, Blue, and Green.  The results are
as follows:

```
o <- c(red = 10, orange = 16, yellow = 24, green = 12)
```

* State a sensible null hypothesis
* State the observed values and the expected values under your null.
* Calculate the Pearson's chi-square statistic `sum((o-e)^2/e)` and show your working; state what its approximate null distribution is
* State the precise definition of p-value and explain what "more extreme" means in this context.
* Calculate the pvalue using `pchisq()` and your result from above, and interpret.
* Excellence question: using random simulation, or otherwise, estimate
  the probability that all four colours have different numbers of
  students.


```{r}
o <- c(red = 10, orange = 16, yellow = 24, green = 12)  # observed
e <- mean(o)                                            # expected
s <- sum((o-e)^2/e)                                     # Pearson's chi-square statistic
## Null: equal probability of choosing each colour
## p-value is the probability, if the null is true, of obtaining the observation
## or an observation more extreme.
## Here "more extreme" means "a larger chi-square statistic"
pchisq(s,df=3,lower.tail=F)                             # pvalue
 ## not significant (exceeds 0.05)
```


```{r}
table(replicate(1000,max(table(table(sample(1:4,62,replace=T))))))
# About 64%
```


# Question 2

A call centre logs the number of support calls received each day for a
  year and tabulates the results:

```{r,include=TRUE}
o <- c(c0=8, c1=12, c2=36, c3=54, c4=67,c5=66, c6=41, c7=37, c8=23, c9=10, c10=11)
o
```

This means that on 8 days there were zero calls, on 12 days there was
one call, on 36 days there were two call, and so on up to 11 days when
there were 10 callouts.  We wish to test the hypothesis that the
number of calls on any day has a Poisson distribution.

* Give a plausible reason why the Poisson distribution might be appropriate

```{asis}
There are a large number $n$ of customers that might call on 
given day, each with a small probability $p$ of calling;
but the product $np$ is moderate.
```

* Verify that the dataset contains 365 observations.  Calculate the
number of calls in the year and then calculate the average number
of calls per day.

```{r}
sum(o)   # number of days
sum(o*(0:10)) # number of callouts
sum(o*(0:10))/sum(o)  # callouts per day
```

* Use your estimated value of the mean number of calls per day as
$\lambda$ in the Poisson distribution to calculate the probability of
having $0,1,2,\ldots,\geqslant 10$ calls on any day.  Remember that
the final probability is "10 calls _or more_" so ensure that
your probabilities sum to one.

```{r}
 1740/365 -> lam
 jj <- dpois(0:9,lam)
 probs <- c(jj,1-sum(jj))
 probs
 sum(probs)   # verification
```

* Use R to calculate the expected number of calls with
$0,1,2,3,4,5,6,7,8,9,\geqslant 10$ in the year.

```{r}
e <- probs*365
e
sum(e)  # verification
```

* Use R to calculate the Badness-of-fit $B=\sum\frac{(o-e)^2}{e}$

```{r}
 B <- sum((o-e)^2/e)
 B
```

* Calculate a $p$-value for your B and interpret (note that the
degrees of freedom is now $11-1-1=9$: there are 11 categories, minus
one because the total is known, minus another one because the
expectation uses an estimated value for $\lambda$.

```{r}
 pchisq(B,df=11-1-1,lower.tail=FALSE)
```

```{asis}
Value exceeds 5\% so not significant.  There is no reason to reject
  the null and the data appear to be Poisson.
```


* excellence question: Someone observes that the number of days with
no calls at all is quite high.  Formulate a sensible null hypothesis
and test it.  Interpret and give a plausible reason for your finding.

```{asis}
a sensible null would be that the number of days with zero calls is
binomial with size 365 and probability
```

```{r}
dpois(0,1740/365)
```

```{asis}
We can provide a p-value as follows
```

```{r}
1-sum(dbinom(0:7,365,dpois(0,1740/365)))
pbinom(7,365,dpois(0,1740/365),lower.tail=FALSE)
```

```{asis}
The pvalue is less than 5\%, so the result is significant! This might
be because phone lines were down or perhaps the data collection system
was flawed.

Note that the null cannot be exactly true if we condition on the total
number of callouts (the probability of having zero zero-call days
(sic) is zero)

It would be (equally?) plausible to say that the number of
zero-call days is Poisson with parameter $365e^{-1740/365}$,
but to see this requires Stein-Chen method (which is only asymptotic,
BTW).
```

```{r}
ppois(7,365*exp(-1740/365),lower.tail=FALSE)
```

```{asis}
The professional would condition on the total number of callouts and
simulate a year using `sample()`

```{r}
table(replicate(1e4,365-length(unique(sort(sample(1:365,1740,replace=TRUE))))))
```

```{asis}
So according to this simulation of ten thousand years, we see how many
years had zero zero-call days, 1 zero-call days, and so on.  The
p-value is then calculated by counting the years with the observed
number, 8, or more:
```

```{r}
n <- 1e4
sum(replicate(n,365-length(unique(sort(sample(1:365,1740,replace=TRUE)))))>=8)/n
sum(replicate(n,365-length(unique(sort(sample(1:365,1740,replace=TRUE)))))>=8)/n
```

```{asis}
which is significant.
```

# Question 3.

This question requires you to use R for random sampling.

Suppose we have a bag with one red, one green, and one blue ball in a
bag.  We make 12 draws with replacement and tally the number of red,
blue, and green balls in our sample.

* How many red, green, and blue balls do you expect to observe?
* For an observation of `c(r=5, g=3, b=4)`, calculate the chi-square
  statistic.  Justify your number of degrees of freedom.
* Using `sample()`, or otherwise, use R to generate a sequence of random draws.
* Generate some sequences of random draws.  For these, calculate: the
  chi-square statistic $B$ and the $p$-value and present histograms or
  tables [you might find `tabulate(...,nbins=3)` useful]
* Excellence question: compare your empirical distribution of $B$
  above with its asymptotic distribution.

```{asis}
Expect four of each.

```{r}
o <- c(r=5, g=3, b=4)
e <- c(4,4,4)  # could use "4"
(B <- sum((o-e)^2/e))
pchisq(B,df=2,lower.tail=FALSE)


tabulate(sample(1:3,12,replace=TRUE),nbins=3)
tabulate(sample(1:3,12,replace=TRUE),nbins=3)

f <- function(o){sum((o-4)^2)/4}
f(tabulate(sample(1:3,12,replace=TRUE),nbins=3))
f(tabulate(sample(1:3,12,replace=TRUE),nbins=3))


replicate(10,f(tabulate(sample(1:3,12,replace=TRUE),nbins=3)))
o <- replicate(1000,f(tabulate(sample(1:3,12,replace=TRUE),nbins=3)))


hist(o,freq=FALSE)
x <- seq(from=0,to=20,len=100)
points(x,dchisq(x,df=2))

plot(ecdf(o))
points(x,pchisq(x,df=2))
```


# Question 4

Give an example of Pearson's chi-square test from your daily life.
Have at least three categories.  State your observation _clearly_,
state your null and expected values _clearly_.  Calculate and state
the value of the chi-square statistic $B$.  State and justify the null
distribution of $B$ (including number of degrees of freedom), give a
$p$-value.  Interpret your findings.

