---
output: pdf_document
title: qqplot and similar techniques
fontsize: 12pt
---


# Introduction

This short handout takes you through `qqnorm()` and related techniques.

Remember that you can type `?qqnorm` at the R prompt to get help
on that command.


# The order statistic

The R system includes a dataset called ``precip` which has information
about precipitation in various US cities.

```{r}
precip <- 
c(170.18, 138.938, 17.78, 123.19, 35.56, 43.688, 52.578, 
33.02, 110.236, 102.108, 98.806, 138.43, 151.892, 122.682, 58.166, 
29.21, 87.376, 89.154, 98.298, 78.232, 77.724, 109.474, 144.272, 
103.632, 106.172, 107.95, 78.74, 80.518, 76.708, 65.786, 124.968, 
93.98, 91.186, 38.1, 76.708, 18.288, 91.948, 115.57, 19.812, 
84.836, 91.694, 102.108, 108.458, 107.95, 41.148, 99.06, 88.9, 
93.98, 79.756, 95.504, 101.346, 91.948, 108.712, 117.856, 62.738, 
124.714, 116.84, 91.186, 19.812, 122.428, 38.608, 82.55, 113.538, 
108.204, 98.552, 44.196, 103.632, 73.914, 37.084, 150.368)
```

You can get help on this dataset by typing ``?precip` at the R
prompt.  We will analyse this dataset.  We will start by a histogram:


```{r}
hist(precip,col='gray')
```

However, observe that the histogram can be misleading.  If we plot a
histogram with a different number of classes, the appearance can be
very different:

```{r}
hist(precip,nclass=50,col='gray')
```

Specifically, the dataset appears to be bimodal.

To plot this dataset in a more standard way, we can just plot the points:

```{r}
plot(precip)
```

However, this form of plot is defective in that the cities are ordered
in the order they appear in the dataset.  Typically, cities would be
ordered alphabetically, and this ordering is not relevant to the
matter at hand (this particular dataset is not ordered
alphabetically).

One common solution is to sort the data, then plot it:

```{r}
plot(sort(precip),pch=16)
grid()
```

we can see straightaway that the dataset is reasonably well-behaved in
that there are no outliers or other abnormalities.  The minimum and
maximum values are not unexpectedly big, and are part of a sequence.

If we want to read off the median, this would be vertical line at $n=35$:


```{r}
plot(sort(precip),pch=16)
abline(v=35)
```

but it would be easier to rescale the horizontal axis from 0 to 1:

```{r}
xprecip <- seq_along(precip)/length(precip)
plot(xprecip,sort(precip),main='vertical line shows 10th percentile')
abline(v=0.1)
```

This form of presentation is known as a *quantile plot*.  
Another form of plot would be to approximate the cumulative
distribution function.  To do this, we switch the x and y axes:


```{r}
plot(sort(precip),xprecip)
```

However, this sort of plot is more conveniently executed using the
`ecdf()` command:

```{r}
plot(ecdf(precip))
```

which gives nice axes etc.  It is always good to see how the technique
works for random data:

```{r} 
plot(ecdf(rnorm(1e5)))
```

# The qqnorm()  command


Looking at the previous figure, if we could ``distort'' the horizontal
axis in such a way as to produce a straight line, that would be a
useful and interesting technique to assess normality.

first we will produce the standard ecdf graph by hand:

```{r}
n <- 1000
x <- sort(rnorm(n))
quant <- seq(from=1/(2*n),to=1-1/(2*n),len=n)
plot(x,quant)
```

Now we will plot the same dataset but distort the x-axis by plotting
instead of the data values, the cumulative distribution function of
the datavalues:

```{r}
quant <- seq(from=0,to=1,len=n)
plot(pnorm(x),quant)
```


See how the ogive has been transformed into a straight line.  It is,
however, more common to distort the y-axis:


```{r}
plot(x,qnorm(quant))
```

See how the mean is given by the intercept and the slope gives the
standard deviation.  It is also possible to use this technique on our
original dataset but the function ``qqnorm()` does much of the hard
work for us:

```{r}
qqnorm(precip)
abline(mean(precip),sd(precip))
```

Above, see how the approximate straight line of the `qqnorm` plot
shows that the data is approximately Gaussian.


# Task

Try the above techniques for the `rivers` dataset, built in to R.  To
get details, type `?rivers` at the R prompt (a "mile" is an obsolete
unit of length equal to about 1.6km, apparently still used in one or
two primitive countries).

# Two-sample nonparametric methods

Given two independent samples, one often wants to see if they are
drawn from the same distribution.  If the two datasets are the same
length, one can simply plot the order statistics of each dataset on
the x- and y- axes respectively:

```{r}
x <- c(0.2, 0.06, 0.73, 0.06, 0.5, 0.31, 0.45, 0.23, 0.42, 0.39, 0.49, 
0.61, 0.53, 0.44, 0.56, 0.07, 0.03, 0.05, 0.39, 0.48, 0.34, 0.34, 
0.02, 0.7, 0.43, 0.38, 0.28, 0.62, 0.22, 0.57)

y <- 
c(0.52, 0.44, 0.56, 0.47, 0.38, 0.67, 0.59, 0.53, 0.36, 0.46, 
0.79, 0.16, 0.39, 0.59, 0.45, 0.64, 0.62, 0.41, 0.27, 0.5, 0.46, 
0.35, 0.58, 0.7, 0.68, 0.54, 0.61, 0.49, 0.29, 0.62)
```

Then we can plot the two ECDFs:


```{r,fig=TRUE}
plot(ecdf(x))
plot(ecdf(y),add=TRUE,col="red")
```

But here it makes sense to plot one against the other, which amounts
to plotting the sorted datasets against one another:

```{r,fig=TRUE}
par(pty="s")
plot(sort(x),sort(y),asp=1,xlim=c(0,1),ylim=c(0,1))
abline(0,1)
```

Function `qqplot()` is a more convenient device; if the datasets are
different lengths, `qqplot()` deals with this appropriately.

```{r,fig=TRUE}
qqplot(x,y)
```


## Task 

Consider the following data, representing yield efficiencies for a
chemical process using two different methods:

```{r}
a <-
c(0.282, 0.303, 0.239, 0.187, 0.131, 0.249, 0.088, 0.368, 0.406, 
0.339, 0.441, 0.257, 0.124, 0.179, 0.334, 0.095, 0.19, 0.175, 
0.275, 0.208, 0.048, 0.082, 0.079, 0.247, 0.443, 0.141, 0.387, 
0.069, 0.067, 0.109, 0.387, 0.254, 0.328, 0.095, 0.452, 0.148, 
0.186, 0.12, 0.088, 0.085, 0.218, 0.176, 0.245)
b <-
c(0.444, 0.436, 0.107, 0.118, 0.414, 0.394, 0.061, 0.608, 0.025, 
0.387, 0.516, 0.403, 0.16, 0, 0.835, 0.226, 0.438, 0.239, 0.686, 
0.169, 0.009, 0.338, 0.356, 0.769, 0.87, 0.001, 0.564, 0.437, 
0.885, 0.422, 0.061, 0.42, 0.515, 0.022, 0.269, 0.211, 0.045, 
0.219, 0.061, 0.507, 0.071, 0.216, 0.333, 0.513, 0.335, 0.565, 
0.08, 0.492, 0.1, 0.237, 0.022, 0.221, 0.502, 0.263, 0.479, 0.476)
```

Use `qqplot()` and `ks.test()` to investigate whether these two
datasets are in fact drawn from the same distribution.  Investigate
whether these datasets are normal.
