---
output: pdf_document
title: Poisson distribution
fontsize: 12pt
---

# Your tasks


Give three or four examples of the Poisson distribution from
your everyday life.  For example, the number of times in a month that
I lose my glasses.  Each day there is a small probability that I will
lose my glasses, but the number of days is about 31, which counts as
large.  Come up with examples from your life.


\begin{itemize}
\item number of people at my bus stop
\item number of times I have to tie my shoelaces in a day
\item number of goals in a football match
\item etc
\end{itemize}

 Using \texttt{rpois()}, verify the wikipedia results for the
mean and variance of the distribution.  Try to verify the formula
given for the median.

Mean is just $\lambda$.

```{r}
mean(rpois(1e6,lambda=3.9))
mean(rpois(1e6,lambda=3.9))
mean(rpois(1e6,lambda=3.9))
```

Median is $\lfloor\lambda+1/3\rfloor$ which is 4 for $\lambda=3.9$:

```{r}
median(rpois(1e6,lambda=3.9))
median(rpois(1e6,lambda=3.9))
median(rpois(1e6,lambda=3.9))
```


 Produce support and likelihood plots for different values of the
observation.  What happens if $n=0$?

Try $n=6$:

```{r}
lam <- seq(from=3,to=10,len=100)
plot(lam,dpois(6,lam))
```


Try $n=0$:

```{r}
lam <- seq(from=3,to=10,len=100)
plot(lam,dpois(0,lam))
```

(just an exponential curve)

Suppose we have an observation of~$n=7$.  Calculate a likelihood
ratio for~$H_1\colon\lambda=6$ against~$H_2\colon\lambda=11$.

```{r}
dpois(7,6)/dpois(7,11)
```

 (harder) Now suppose you are working with $L=\log\lambda$.  Plot
a support curve for $L$.


```{r}
lam <- seq(from=3,to=16,len=100)
plot(lam,dpois(7,lam,log=TRUE))
```

 (also hard).  Now bring in Bayesian logic.  Suppose
  Your\footnote{It is conventional and polite to capitalize ``You''
    and ``Your'' when using Bayesian logic.} prior for $L$ is Gaussian
  with mean zero and standard deviation 1; a single observation of
  $n=3$ is made.  Plot a support curve for $L$; superimpose support
  curves for different priors.
  
  
```{r}
lam <- seq(from=0.1,to=10,len=100)
L <- log(lam)
plot(L,dnorm(L,log=TRUE))
plot(L,dnorm(L) + dpois(6,exp(L)))
```


 Suppose we test a null hypothesis $H_0\colon\lambda=1$, against
an alternative~$H_A\colon\lambda>1$.  State the precise definition of
p-value, then find a value of $n$ such that a test (of the form
``reject $H_0$ if the observation is greater than
$n$") has size less than 5\%.


pvalue is the probability, if the null is true, of obtaining the
observation or an observation more extreme.

```{r}
sum(dpois(0:3,1))
sum(dpois(0:2,1))
```

reject null if $n\geq 3$ (2 is not big enough, size is over 0.05)



 Now consider a specific alternative hypothesis, specifically
$H_A\colon\lambda=\lambda_0$.  Find the minimum value of $\lambda_0$
so that $\alpha\leqslant 0.05$ and $\beta\leqslant 0.2$.


Need probability of correctly rejecting null to be over 0.8:

```{r}
1-sum(dpois(0:2,lambda=4.2))
1-sum(dpois(0:2,lambda=4.3))
```


Value between 4.2 and 4.3

