---
output: html_document
title: logistic regression suggested analysis
fontsize: 12pt
---


# STAT601 lab, week 11, suggested analysis

## Task 1

Consider the following dataset:

`y <- c(F,F,T,F,F,T,T,T,F,F,T,T,T,F,T,T)`.

These observations represent success or failure of a person trying to
run a particular distance within a prescribed time on successive days.
It is reasonable to assume that this person gets fitter with time.
Analyse this data using a logistic regression and justify your use of
one-sided or two-sided tests.

```{r}
logistic <- function(x){exp(x)/(1+exp(x))}
LO <- function(p){log(p/(1-p))}
y <- c(F,F,T,F,F,T,T,T,F,F,T,T,T,F,T,T)
x <- seq_along(y)
plot(x,y,pch=16)
fit <- glm(y~x,family='binomial')
jj <- fit$coefficients
points(x,logistic(jj[1] + jj[2]*x),type="l",col="red")
summary(fit)
```

Note that the pvalue is half that reported by R: `glm()` gives
two-sided pvalues by default, and we need a one-sided test because the
subject is getting fitter.  The pvalue for the one-sided test will be
about 0.066, not significant.


## Task 2

Download the `e3mg` dataset from last week.  We are interested in the
length of an economic recession being greater than five years.  Create
a Bernoulli RV with `1` representing a recession being longer than
five years and `0` meaning the recession was less than five years.
Analyse this using `glm()`.

First we need some data wrangling:

```{r}
a <- read.table("e3mg.txt")
a <- cbind(a,long=as.numeric(a$rec_len>5)) 
head(a)
```

And we must not forget basic EDA:


```{r}
table(a$long)
```

[this is a less than ideal split, it would have been better to have a
50-50 split (such a dataset is called "balanced"), but here as often
in life we have to deal with what we have, not what we would like].

We will try a visual plot:

```{r}
plot(long~oil.price,data=a)
```

Note that the points are overplotted, we could use `jitter()` to
distinguish them:

```{r}
plot(jitter(long,amount=0.1)~oil.price,data=a)
fit <- glm(long~oil.price, data=a,family='binomial')
summary(fit)
points(a$oil.price,fitted.values(fit),type="p",col='red')
```

(we can do the other variables in a similar way)


```{r}
summary(glm(long~oil.price+direct.tax+interest.rate+saving.ratio+investment,data=a))
```

I would suggest dropping `interest.rate`:

```{r}
summary(glm(long~oil.price+direct.tax+saving.ratio+investment,data=a))
```



* Apply `glm()` to a Bernoulli column of your portfolio dataset.  
