---
output: html_document
title: "Type I and type II errors with Gaussian distributions"
fontsize: 12pt
---

# Your first task

Vary the number of observations (choose three or four representative
values in the range 5-50), and find the power of the test against for
the same null, $H_0\colon\mu=0$ against different alternatives:
$H_A\colon\mu=0.1,0.2,0.5,1,2$.  Use `pnorm()` etc and then verify
numerically.


```{r}
f <- function(mu1,n){pnorm(qnorm(0.95,sd=1/sqrt(n)),mean=mu1,sd=1/sqrt(n),lower.tail=FALSE)}
m1 <- c(0.1,0.2,0.5,1,2)
plot(m1,sapply(m1,f,n=30),type="b",ylab="power",main="n=30")
plot(m1,sapply(m1,f,n=50),type="b",ylab="power",main="n=50")
```


Numerically:

```{r}
f(0.2,30)
n <- 30
table(rnorm(1e6,mean=0.2,sd=1/sqrt(n)) > qnorm(0.95,sd=1/sqrt(n)))
```

# Tradeoff between type I and type II errors.

Suppose we have $n=5$ observations and wish to test $H_0\colon\mu=0$
against $H_A\colon\mu=1$.  Here we use one-sided tests for simplicity.

```{r}
p <- 0.05 # pvalue
u <- qnorm(1-p,0,sd=1/sqrt(5))  # u is the test: reject H0 if xbar > u
pnorm(u,1,sd=1/sqrt(5),lower.tail=FALSE)  # power of the test
```

The above calculates directly the power of the test for $\alpha=0.05$

# Task

* For $H_0\colon\mu=0, H_A\colon\mu=1$.  Keeping $n=5$ for the moment,
vary the size $\alpha$ of the test and investigate how the power
changes in response.

```{r}
n <- 5
f <- function(alpha){pnorm(qnorm(1-alpha,sd=1/sqrt(n)),mean=1,sd=1/sqrt(n),lower.tail=FALSE)}
alpha <- seq(from=0.01,by=0.01,to=0.1)
plot(alpha,sapply(alpha,f),type='b')
```

* For $H_0\colon\mu=0, H_A\colon\mu=1$.  Vary $n$ and see how the power
  changes, for a fixed value of $\alpha$, in response.  Try $n=8$,
  $n=15$, $n=20$ for a start.  Describe the effect of increasing $n$
  on the power of the test.
  
```{r} 
f <- function(n){pnorm(qnorm(1-0.05,sd=1/sqrt(n)),mean=1,sd=1/sqrt(n),lower.tail=FALSE)}
n <- c(8,15,20)
plot(n,sapply(n,f),type='b',ylab="power")
```
  

* (harder) Consider a different alternative hypothesis, say
  $H_A\colon\mu=0.5$.  How do the answers to the two questions above
  change when we consider this new $H_A$?  How about
  $H_A\colon\mu=0.1$ or $H_A\colon\mu=0.01$?

```{r}
f <- function(n,mu1){pnorm(qnorm(1-0.05,sd=1/sqrt(n)),mean=mu1,sd=1/sqrt(n),lower.tail=FALSE)}
n <- c(8,15,20,50,100)
plot(n,sapply(n,f,mu1=0.5),type='b',ylab="power")
plot(n,sapply(n,f,mu1=0.01),type='b',ylab="power")
```


# Student t test

```{r}
hist(replicate(10000,sd(rnorm(n))))
```

Try this yourself with different values of $n$.

```{r}
hist(replicate(10000,sd(rnorm(3))))
```

# Task

(in all the tasks below, use numerical simulation)

* With $n=6$, what is the probability of overestimating the standard
  deviation by a factor of 2 or more?  What is the probability of
  underestimating by a factor of 2 or more? What is the probability of
  estimating the sd correct to within a factor of 2? (use numerical
  simulation)
  
```{r}
table(replicate(1000,sd(rnorm(6))>2))
```
  
  
* What are the corresponding probabilities if $n=10$ and $n=15$?

  
```{r}
table(replicate(1000,sd(rnorm(10))>2))
table(replicate(1000,sd(rnorm(15))>2))
```

* (harder) How many observations do we need to make to ensure that the
  probability of overestimating or underestimating by a factor of 2 or
  more is less than 5\%? (use numerical simulation).

```{r}
OK <- function(n){
  est <-sd(rnorm(n))
   (est>0.5) & (est<2)
}
c(
n5=sum(replicate(10000,OK(5))),
n6=sum(replicate(10000,OK(6))),
n7=sum(replicate(10000,OK(7))),
n8=sum(replicate(10000,OK(8))),
n9=sum(replicate(10000,OK(9)))
)
```

(looks like you need $n\geq 7$)

*  With $n=4$, what is the expected value of the estimated variance?
  [hint: `var(x)` gives the estimated variance of `x`] 
  
```{r}
  mean(replicate(100000,var(rnorm(4))))
```

* Again with $n=4$, what is the expected value of the estimated
  standard deviation?  [this is not a trick question but can be
  confusing.  Do not stress].


```{r}
mean(replicate(1e5,sd(rnorm(4))))
```


# Your task
 
Calculate the power of the t-test, varying the number of observations
from $n=5,10,20$ and the alternative hypothesis
$\mu=0.1,0.2,0.5,1.0,2$.
 

```{r}
power <- function(n,mu){
  sum(replicate(1000,t.test(rnorm(n),alternative="less",mu=mu)$p.value<0.05))/1000
}
power(10,0.2)
power(10,0.2)
power(10,1.0)
o <- expand.grid(n=c(5,10,20),mu=c(0.1,0.2,0.5,1,2))
cbind(o,power=apply(o,1,function(x){power(x[1],x[2])}))
```

