---
output: html_document
title: Pearson's chi square test _suggested analyis_
fontsize: 12pt
---


# Task 1: chi-square analysis for favourite colours

A researcher asks 89 students which of four colours is their
favourite.  The observations are:


```{r}
o <- c(red=10,green=16,blue=17,yellow=20)
```

State a sensible null and analyse it using Pearson's chi-square test.


## suggested analysis

all colours have equal probability (0.25).

```{r}
e <- o
e[] <- mean(e)
e
B <- sum((o-e)^2/e)
pchisq(B,df=3,lower.tail=FALSE)
```

p-value exceeds 0.05, fail to reject null


### Task 2: chi-square for Poisson observations.

The number of goals scored by the home team in 47 soccer games played
in the UK in one week is as follows:


```{r}
o <- c(6,9,15,11,3,2,1)
names(o) <- paste("goals",0:6,sep="")
o
```

Remembering that counts should be $>=5$, test the null that the number
of goals is in fact Poisson.  What would be a good justification for
the counts being Poisson?

### suggested analysis

A soccer match has a large number of minutes during each of which
there is a small probability of scoring (you may ignore
nonindependence between successive minutes).

```{r}
(lambda <- sum(o*(0:6))/sum(o))  # goals / matches = mean
p <-dpois(0:3,lambda)
p <- c(p,1-sum(p))  # calculate prob(X=-0,1,2,3, >= 4)
(e <- p*sum(o))  # expected values
o <- c(6,9,15,11,6)
B <- sum((o-e)^2/e)
pchisq(B,df=5-1-1,lower.tail=F) # subtract one for total, one for estimated lambda
```

Large pvalue, fail to reject null.



## Task 3: chi-square for family gender distribution.


A sample of 400 families with exactly 5 children is surveyed and the
number of boys is recorded (the number of boys is in the range 0-5).
The results are as follows:


```{r}
o <- c(17, 70, 143, 117, 47, 6)
names(o) <- paste("boys",0:5,sep="")
o
```

* Test the null that each birth is a boy with probability 0.5


```{r}
e <- sum(o)*dbinom(0:5,5,0.5)
B <- sum((o-e)^2/e)
pchisq(B,df=6-1,lower.tail=F) # no estimation here
```

Reject null.


* Estimate the probability of a birth being a boy and test the null
    that the number of boys in each family is binomial

### Suggested analysis


```{r}
(p <- sum(o*(0:5))/(sum(o)*5))  # number of boys/number of children = #boys/#(families*5)
e <- sum(o)*dbinom(0:5,5,p)
B <- sum((o-e)^2/e)
pchisq(B,df=6-1-1,lower.tail=F) # estimate reduces DF by 1
```

Fail to reject null, observations consistent with a binomial but not
with probability one half.


### Task 4: chi-square as an alternative to Fisher's exact test

the following contingency table shows the results of American and Russian
(USSR) missions to Mars.

```{r}
M <- matrix(c(16,5,6,16),byrow=TRUE,2,2)
dimnames(M) <- list(nation=c("USSR","USA"),mission=c("success","fail"))
M
```


* Given the marginal totals, calculate expected values for each of the
  four table entries
* Calculate a badness value
* Using one degree of freedom, calculate a p-value and compare with
  the result from Fisher's exact test.

### Suggested analysis

Null would be that the two agencies have the same probability of
success.

```{r}
(e <- outer(rowSums(M),colSums(M))/sum(M))  # expectation
B <- sum((M-e)^2/e)   # badness
pchisq(B,df=1,lower.tail=FALSE)  ## Pearson's chi-square pvalue
fisher.test(M)$p.value
```


Pearson's chi-square questionable because of small entries; Fisher is
preferable here.  Pearson gives a pvalue about half that of Fisher
(which I found surprising: Fisher's test is more powerful [in the
sense of having a smaller $\beta$ ( that is, a smaller probability of
committing a type-two error)] than Pearson which usually translates
into having a smaller pvalue...but not in this case).

### Task 5: binomial vs hypergeometric

A total of 400 random observations are drawn from _either_ a binomial
distribution $\operatorname{Bin}(4,0.5)$ ($H_B$), _or_ a
hypergeometric distribution with $m=n=k=4$, ($H_H$), but it is not
known which.  Here is a table of the observations:

```
  0   1   2   3   4 
  7  70 222  93   8
```

Thus, "0" was observed 7 times, "1" was observed 70 times, and so on
up to "4", which was observed 8 times.

* Conduct a Pearson's chi-square test on the hypothesis of binomial
   distribution
* Conduct a Pearson's chi-square test on the hypothesis of hypergeometric
   distribution
* Calculate a likelihood ratio for these two hypotheses
* Bayesian: suppose the priors are $\operatorname{Prob}(H_H)=10^{-3}, \operatorname{Prob}(H_B)=0.999$.  Calculate the posterior probabilities.

### Suggested analysis

```{r}
o <- c(7,  70, 222,  93,   8)
supp_B <- sum(o*dbinom(0:4,4,0.5,log=TRUE))
supp_H <- sum(o*dhyper(0:4,4,4,4,log=TRUE))
supp_B-supp_H
exp(supp_B-supp_H)
```

Likelihood ratio strongly supports hypergeometric over binomial.

First do the binomial chi-square test:

```{r}
e <- sum(o)*dbinom(0:4,4,1/2)
pchisq(sum((o-e)^2/e),df=4,lower.tail=FALSE)
```

Next the Hypergeometric chi-square:

```{r}
e <- sum(o)*dhyper(0:4,4,4,4)
pchisq(sum((o-e)^2/e),df=4,lower.tail=FALSE)
```

reject binomial, fail to reject hypergeometric.
Now Bayes:

```{r}
prior_H <- 1e-3
prior_B <- 1-prior_H
a <- prior_H * exp(supp_H)
b <- prior_B * exp(supp_B)
posterior_H <- a/(a+b)
posterior_B <- b/(a+b)
c(hypergeometric=posterior_H,binomial=posterior_B)
```

The small prior for $P_B$ is overwhelmed by data
