---
title: "MATPMDB lab, week 1, binomial distribution"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE) # change to TRUE when rendering solutions; this changes
                                       # the default visibility (which is overridden by explicitly
                                       # setting  include=TRUE in chunk options)
```


\newcommand{\bling}{XXXxxxXXxX}

This tutorial focusses on the binomial distribution.  This tutorial
builds on chapter 1 of the course manual.  You will need
to have R installed on your computer and be able to enter commands and
interpret its output.


### Question

Let $X\sim\operatorname{bin}(13,0.8)$.  This means that we have 13 independent
  Bernoulli trials, each with a probability of 0.8 of success.

Compute the following probabilities using the R function
`dbinom(k,n,p)` and the formula: $P(X=k) =
\binom{n}{k}p^k(1-p)^{n-k}$.  Note that $\binom{n}{k} =
\frac{n!}{k!(n-k)!}$.  Check that you get the same answer using both
methods.

* $P(X = 10)$
```{r}
factorial(13)/(factorial(10)*factorial(13-10))*0.8^10*(1-0.8)^(13-10)

#alternatively
choose(13,10)*0.8^10*0.2^3

# better:

dbinom(10,13,0.8)
```


```{asis}
It is often useful to define variables and use these variables in the
functions.
```

```{r}
n <- 16
p <- 0.8
x <- 13
dbinom(x, n, p)
```

* $P(X \leq 9)$

```{asis}
there are a number of techniques we can use here
```

```{r}

dbinom(0,13,0.8) + dbinom(1,13,0.8) + dbinom(2,13,0.8) +
dbinom(3,13,0.8) + dbinom(4,13,0.8) + dbinom(5,13,0.8) +
dbinom(6,13,0.8) + dbinom(7,13,0.8) + dbinom(8,13,0.8) +
dbinom(9,13,0.8)


sum(dbinom(0:9,13,0.8))

pbinom(9,13,0.8)
```
```{asis}
(see how all the different methods give the same result).
```

### Question

According to one study, The proportion of adults (18 years or older)
who admit to texting while driving is 47\%. Suppose you randomly
select ten adult drivers and ask if they text while driving (do you
think this percentage is realistic?); you may assume the drivers are
independent.  Say $X$ is the number of drivers in the sample who admit
to texting while driving.  If $X$ is modelled by a binomial random
variable, state $n$ and $p$.

```{asis}
$X \sim\operatorname{bin}(10,0.47)$.
```

* Compute the probability mass function for all values of $X$ using R.

```{r}
n <- 10
x <- 0:10
p <- 0.47
y <- dbinom(x, n, p)
names(y) <-  x
y

# check that probabilities sum to 1:
sum(y)
```

* Construct a probability histogram of $X$.

```{r}

plot(x,dbinom(x,n,p),type='h')
```

* Compute the expected value and standard deviation for $X$ and verify numerically

```{r}
n*p  # mean
sqrt(n*p*(1-p))  # standard deviation
mean(rbinom(1e6, n, p))  # numerical simulation of mean
sd(rbinom(1e6, n, p))    # numerical simulation of sd
```

### Question

We take a fair coin (that is, one with probability $p$ of landing
heads = 0.5) and toss it $n=1000$ times.  

* Plot the probability of observing a total of $r=400,401,\ldots, 600$
heads.

```{r}
x <- 400:600
plot(x,dbinom(x,1000,1/2))
```

* Repeat the simulation experiment in the previous question using
different values of $p$.  

```{r}
f <- function(p){dbinom(0:1000,1000,p)}
x <- 0:1000
plot(x,f(0.5),type="l")
points(x,f(0.6),type="l",col="red")
points(x,f(0.7),type="l",col="orange")
points(x,f(0.8),type="l",col="yellow")
points(x,f(0.14),type="l",col="green")
```

### Question

A test contains 10 questions. Each question is multi-choice with four
options.  Let $X$ be the number of questions that a student answers
correctly.  Suppose that the student has not studied, so their
strategy is to randomly guess the answers.  This means that the
probability of guessing the correct answer is $p=1/4$.

* What distribution can be used to model the number of correct answers?

```{asis}
Binomial distribution, with $n=10$ and  $p=0.25$.
```

* What is the probability that the student answers exactly 5 questions correctly?

```{r}
dbinom(5, 10, 0.25)
```

* The pass mark for this test is to score 6 or more correct answers.
  What is the probability that our student passess the exam?

```{r}
sum(dbinom(6:10,10,1/4))
pbinom(5,10,1/4,lower.tail=FALSE)
```

```{asis}
A little under 2\%
```

### Question

(harder) Given an integer $n>0$, and a random variable
$X\sim\operatorname{bin}\left(2n,\frac{1}{2}\right)$, we are interested
in the probability $p_n$ that the number of failures is equal to the
number of successes, that is, $\operatorname{Prob}(X=n)$.

* Use `dbinom()` to calculate $\operatorname{Prob}(X=n)$ for different
  values of $n$.  Does the probability increase or decrease with
  increasing $n$?

```{r}
n <- 1:50
p <- sapply(n,function(n){dbinom(n,2*n,1/2)})
plot(n,p)
plot(n,p,log="xy")
```

* Google for "Stirling's formula".  Stirling's formula states that,
for large values of $n$, \[ n!\approx\sqrt{(2\pi n)}e^{-n}n^n.\]
Stirling's formula is only approximate, but is very much easier to
deal with than exact factorials.  The approximation gets better as $n$
gets larger.

* Calculate the percentage error in using Stirling's approximation for
  factorial $n$, for a range of values of $n$

```{r}
stirling <- function(n){sqrt(2*pi*n)*n^n/exp(n)}
stirling(10)
factorial(10)
error <- function(n){abs(stirling(n)-factorial(n))/factorial(n)}
plot(n,error(n))
plot(n,error(n),log="y")
```

* Obtain an algebraic expression for $\operatorname{Prob}(X=n)$ as
  above (hint: lots of terms cancel out very nicely).

```{asis}
we would have 
\[
\binom{2n}{n\,\, n}\left(\frac{1}{2}\right)^n\left(\frac{1}{2}\right)^n=\frac{(2n)!}{(n!)^24^{2n}}\]

and after some calculation we would have
$\operatorname{Prob}(X=n)=\frac{1}{\sqrt{\pi n}}$.
```

* Assess the accuracy of Stirling's approximation in this context

```{r}
error <- function(n){
   exact <- dbinom(n,2*n,1/2)
   approx <- 1/sqrt(pi*n)
   abs(exact-approx)/exact
}
plot(n,sapply(n,error),log='xy')
```

