---
output: html_document
title: Fisher's exact test
fontsize: 12pt
---



\newcommand{\prob}[1]{\operatorname{Prob}\left(#1\right)}

Remember that you can type \code{?fisher.test} at the R prompt to get
help on Fisher's test.  You are also encouraged to search for
``Fisher's exact test'' on Google; the wikipedia page is an excellent
source of information.

There are many ways to analyse two-by-two contingency tables, of which
Fisher's test is the most commonly used.  Your lecturer has a
long-standing and intense love-hate relationship with Fisher's exact
test and has written many research articles contributing to the
debate.  


# Example 1

Consider an experiment in which a number of students are classified
into male or female, and left- or right- handed.  We are interested in
whether there is a gender difference in handedness.

The data might look like this


```{r} 
a <- read.table("handedness.txt",header=TRUE)
head(a)
```

There is a  file called `handedness.txt` on AUTonline.  We can tabulate as follows:

```{r}
contingency <- table(a$gender,a$handedness)
```

Often in lectures you will see me improving the appearance of the output with the following cosmetic improvement:

```{r}
dimnames(contingency) <- list(gender=c("M","F"),handedness=c("L","R"))
contingency
```

This makes no difference to the statistics, it just looks nicer.

## your tasks

* What is the probability of a person in this dataset being Female?
  What is the probability of a person being Left-handed?
  What is the probability of a left handed person being male?
  What is the probability of a female being right-handed?
* State a sensible null hypothesis  
* Translate `contingency` into the white ball, black ball language
  used in the R help page under `?hypergeometric`
* Use `dhyper()` to calculate the probability of observing
  `contingency` with given marginal totals.
* Write out all possible contingency tables with the same marginal
  totals.  That is, all two-by-two tables with non-negative integer
  entries that have the same rowsums and columnsums as table
  `contingency` in the R code above.
* Define "more extreme" to mean "contingency table with a smaller
  `dhyper()` probability", state the precise definition of p-value,
  and calculate it.  Is this a one-sided test or a two-sided test?
* Verify with `fisher.test()`.

\newpage


# Example 2


Consider the following contingency table:

```{r}
M <- matrix(c(7,1,2,5),2,2)
dimnames(M) <- list(professor=c(T,F),owns_car=c(T,F))
M
```

This shows incidence of car ownership among AUT academic staff.  One
might expect a professor, earning a massive salary, to be more likely
to be able to afford a car than a mere Senior Lecturer earning just
above minimum wage.

## your tasks

* State a sensible null hypothesis and specify whether a one-sided or
  two-sided test is more appropriate.
* Use `fisher.test()` to test your null and interpret.
* Verify using `dhyper()`.


\newpage

# Example 3

Nine English and seven Scottish cows are tested for a particular
disease and six cows, of which four are Scottish, test positive.

## Your tasks

* Translate this into a contingency table
* State a sensible null and decide whether one-sided or two-sided tests are appropriate
* Test your null using `fisher.test()` and interpret.


# Example 4

Technically, Fisher's test applies only to two-by-two contingency
tables but is easily generalized to larger tables.  Enumerating the
sample space (that is, tables with matching marginal totals) is harder
than the two-by-two case, but easy with computers.  The null
distribution is not hypergeometric but again easy for computers to
evaluate.  In R, function `fisher.test()` takes contingency tables of
any size.  


The following table presents data on students' favourite
colours:


```{r}
M <- matrix(c(5,8,3,2,2,6,1,2,8),3,3)
dimnames(M) <- list(paper=c("ENGE502","STAT601","STAT805"),colour=c("red","blue","green"))
M
```

The following table presents data on a particular family in Auckland,
and how often they cooked different types of food over the last few
weeks.


```{r}
M <- matrix(c(1, 3, 5, 3, 6, 7, 5, 2, 2, 4, 2, 6, 2, 12, 3, 9, 4, 6, 0, 5),5,4)
dimnames(M) <- list(person=c("robin","lesley","alice","zac","annabel"),colour=c("pasta","soup","pizza","salad"))
M
```



## Your tasks

state sensible nulls for these two datasets and analyse using Fisher's
exact test.  Hint: you may need to use argument `simulate.p.value` to
enable R to operate (why do you think this is needed?)




