---
output: html_document
title: qqplot and similar techniques, suggested analysis
fontsize: 12pt
---


# Your tasks

## Task 1

Try the above techniques for the `rivers` dataset, built in to R.  To
get details, type `?rivers` at the R prompt (a "mile" is an obsolete
unit of length equal to about 1.6km, apparently still used in one or
two primitive countries).

**Just load the data and look at using different techniques**

```{r}
rivers
hist(rivers)
plot(ecdf(rivers))
qqnorm(rivers)
abline(mean(rivers),sd(rivers))
```


**All the techniques point to a strong right skew; take a log transform:**

```{r}
lr <- log(rivers)
hist(lr)
plot(ecdf(lr))
qqnorm(lr)
abline(mean(lr),sd(lr))
```

**Still not entirely normal but better than the original data**

## Task 2

Consider the following data, representing yield efficiencies for a
chemical process using two different methods:

```{r}
a <-
c(0.282, 0.303, 0.239, 0.187, 0.131, 0.249, 0.088, 0.368, 0.406, 
0.339, 0.441, 0.257, 0.124, 0.179, 0.334, 0.095, 0.19, 0.175, 
0.275, 0.208, 0.048, 0.082, 0.079, 0.247, 0.443, 0.141, 0.387, 
0.069, 0.067, 0.109, 0.387, 0.254, 0.328, 0.095, 0.452, 0.148, 
0.186, 0.12, 0.088, 0.085, 0.218, 0.176, 0.245)
b <-
c(0.444, 0.436, 0.107, 0.118, 0.414, 0.394, 0.061, 0.608, 0.025, 
0.387, 0.516, 0.403, 0.16, 0, 0.835, 0.226, 0.438, 0.239, 0.686, 
0.169, 0.009, 0.338, 0.356, 0.769, 0.87, 0.001, 0.564, 0.437, 
0.885, 0.422, 0.061, 0.42, 0.515, 0.022, 0.269, 0.211, 0.045, 
0.219, 0.061, 0.507, 0.071, 0.216, 0.333, 0.513, 0.335, 0.565, 
0.08, 0.492, 0.1, 0.237, 0.022, 0.221, 0.502, 0.263, 0.479, 0.476)
```

Use `qqplot()` and `ks.test()` to investigate whether these two
datasets are in fact drawn from the same distribution.  Investigate
whether these datasets are normal.

**Just do it:**


```{r}
qqnorm(a,pch=16)
jj <- qqnorm(b,plot.it=FALSE)
points(jj$x,jj$y,pch=16,col='red')
```

**Highly non-normal, t-test not applicable here**

```{r}
plot(ecdf(a))
plot(ecdf(b),add=TRUE,col="red")
legend("topleft",pch=16,lty=1,col=c("black","red"),legend=c("a","b"))
abline(v=0.33)
```

**The ecdf() plot suggests that $D\simeq 0.35$ (vertical line drawn on
 the graph at maximal discrepancy point).  Verify with Kolmogorov
 test:**

```{r}
ks.test(a,b)
```

**p-value given by Kolmogorov-Smirnov is highly significant, test does
  not rely on the assumption of normality**

**Try a qqplot:**

```{r}
qqplot(a,b)
abline(0,1)
```