---
output: pdf_document
title: Poisson distribution
fontsize: 12pt
---

# The Poisson distribution


This short handout takes you through some theory and practical work
with the Poisson distribution, and has some exercises at the end.

  
Remember that you can type \texttt{?rpois} at the R prompt to get help
  on the Poisson distribution.  You are also encouraged to search for
  ``Poisson distribution'' on Google; the wikipedia page is an
  excellent source of information.  }


\section{The Poisson Distribution}

You will recall the probability mass function of the Poisson from
lectures:

$$
\newcommand{\prob}[1]{\operatorname{Prob}\left(#1\right)}
\prob{X=n} =\frac{e^{-\lambda}\lambda^n}{n!}
$$

The R commands are \texttt{rpois} etc, so type \texttt{?rpois} for help.

Try to think of a few examples from your day-to-day life that follow
the Poisson distribution.  Remember that the Poisson is the limit of
the binomial distribution as $n$ goes to infinity, $p$ goes to zero,
but maintaining $np$ constant.  So your examples should have large $n$
and small $p$.

We will just do a few things first with \texttt{lambda = 2.1}:                  
                  
```{r}
f <- function(n,lambda){exp(-lambda)*lambda^n/factorial(n)}
f(0:10,lambda=2.1)
dpois(0:10,lambda=2.1)

f(0:10,2.1)-dpois(0:10,lambda=2.1)
```

See how close the two methods are.  Try to assess the difference
between the two methods, which should be small; can you quantify it?
Produce a graphical representation of the difference between the two methods.
Here is one method:
```{r}
plot(dpois(0:10,lambda=2.1),f(0:10,2.1))
```

but note that this is a very poor graph.  Try to improve it! We will use \texttt{dpois()} from now on as it is more accurate and faster.  One of the first things to do is to check that the sum of the
probabilities is one, for a few different values of $\lambda$:

```{r}
sum(dpois(0:10,lambda=0.4))
sum(dpois(0:10,lambda=1.2))
sum(dpois(0:10,lambda=3.3))
sum(dpois(0:10,lambda=3.8))
```

(actually this will be very very slightly less than one (why?))

See the following figure:

```{r}
n <- 0:10
plot(n,dpois(n,lambda=2.3),type="h",lwd=5)
```

We can also try a likelihood approach.  Suppose we are trying to
estimate $\lambda$ and have an observation of $n=7$.  What values of
$\lambda$ are consistent with this?  (see Figure~\ref{like}).

```{r}
lambda <- seq(from=2,to=15,len=100)
like  <- dpois(7,lambda=lambda)
plot(lambda,like)
abline(v=7)
```


```{r}
lambda <- seq(from=2,to=15,len=100)
support  <- dpois(7,lambda=lambda,log=TRUE)
plot(lambda,support-max(support))
abline(v=7)
abline(h=c(0,-2))
```

# Your tasks

\begin{itemize}

\item Give three or four examples of the Poisson distribution from your everyday life.  For example, the number of times in a month that I lose my glasses.  Each day there is a small probability that I will lose my glasses, but the number of days is about 31, which counts as large.  Come up with examples from your life.

\item Using \texttt{rpois()}, verify the wikipedia results for the
mean and variance of the distribution.  Try to verify the formula
given for the median.

\item Produce support and likelihood plots for different values of
the observation.  What happens if $n=0$?

\item Suppose we have an observation of~$n=7$.  Calculate a likelihood
ratio for~$H_1\colon\lambda=6$ against~$H_2\colon\lambda=11$.

\item (harder) Now suppose you are working with $L=\log\lambda$.  Plot
a support curve for $L$.

\item (also hard).  Now bring in Bayesian logic.  Suppose
  Your\footnote{It is conventional and polite to capitalize ``You''
    and ``Your'' when using Bayesian logic.} prior for $L$ is Gaussian
  with mean zero and standard deviation 1; a single observation of
  $n=3$ is made.  Plot a support curve for $L$; superimpose support
  curves for different priors.

\item Suppose we test a null hypothesis $H_0\colon\lambda=1$, against
an alternative~$H_A\colon\lambda>1$.  State the precise definition of
p-value, then find a value of $n$ such that a test (of the form
``reject $H_0$ if the observation is greater than
$n$") has size less than 5\%.

\item Now consider a specific alternative hypothesis, specifically
$H_A\colon\lambda=\lambda_0$.  Find the minimum value of $\lambda_0$
so that $\alpha\leqslant 0.05$ and $\beta\leqslant 0.2$.
\end{itemize}

