---
output: pdf_document
title: Linear regression
fontsize: 12pt
---


# introduction

This short handout takes you through linear regression.  Go through
the material, typing the R commands at the R prompt, making sure you
understand the content.  There are some tasks at the end.
       
Remember that you can type `?lm` at the R prompt to get help
on linear regression.

# Linear regression: trees.

The R system includes a dataset called `trees`:

```{r}
data(trees)
head(trees)
```

We will use linear regression to analyse this dataset.  We will start
by analysing the dependency of volume on tree height; The first step
is to plot a scattergraph:


```{r}
plot(Volume~Height,data=trees)
```

The points are reasonably scattered, without any outliers or
nonlinearity.  The points are showing broadly uniform variance about a
line.  We can use R to calculate the best-fit line using the
`lm()` function:


```{r}
lm(Volume~Height,data=trees)
```

This shows that the relation is approximately

\[
V =  -87.124    +    1.543H
\]

We can get more information by using the `summary()` command:


```{r}
fit <- lm(Volume~Height,data=trees)
summary(fit)
```

This gives more information such as p-values and standard errors. For
example, the slope is 1.5433 (estimated) but with an error of $\pm
0.3839$.


We can plot the regression line on the scattergraph as follows.

```{r}
plot(Volume~Height,data=trees)
abline(fit)
```

However, this model has a serious defect in that short trees are
predicted to have negative volumes, an absurdity.  We can mitigate
this defect by using a logarithmic transform:


```{r}
plot(Volume~Height, data=trees, log='xy')
```

The `plot()` command takes a `log` argument, in this case
instructing R to use logarithmic transforms on the x and y axes.

However, `abline()` does not work with this type of plot, so we will
have to work manually:


```{r}
plot(log(Volume)~log(Height),data=trees)
logfit <- lm(log(Volume)~log(Height),data=trees)
abline(logfit)
```

And to see the details we use `summary()`:

```{r}
summary(logfit)
```


## Using Girth instead

The girth of the tree is a more useful predictor than height:

```{r}
plot(Volume~Girth,data=trees,log='xy')
```

We can try linear regression:


```{r}
logfit2 <- lm(log(Volume)~log(Girth),data=trees)
summary(logfit2)
```

The plot is more convincing than the numbers, IMHO:

```{r}
plot(log(Volume)~log(Girth),data=trees)
abline(logfit2)
```


It is possible to use both height and weight together as regressors:

```{r}
summary(lm(Volume ~ Girth + Height, data=trees))
```

This would mean that $V = -57.9877 + 4.7082G + 0.3393H$.  Again the
model has a defect that small trees are predicted to have negative
volume.  So it might be better to use logarithmic transformation:

```{r}
summary(lm(log(Volume) ~ log(Girth) + log(Height), data=trees))
```

This would mean that $\log V = -6.63 + 1.98\log G + 1.117\log H$,
or $V = e^{-6.63}G^{1.98}H^{1.117}$.

If the height and weight have combined effects we can use the star
operator:


```{r}
summary(lm(log(Volume) ~ log(Girth) * log(Height), data=trees))
```

This would indicate that $\log V = -3.68 + 0.79\log G + 0.43\log H +
0.27\log G*\log H$.  Do you think this is a better model?



# Your Tasks

## Task 1:  Height-weight analysis

Download file `heightweight.txt` from AUTonline and load it into R, using

`a <- read.table(file.choose(),header=TRUE)`

Then use commands such as 

`plot(weight~height,data=a)`

to produce a scattergraphs.  

What is the predicted weight of someone with a height of 2m exactly?
What is the predicted weight of someone with a height of 0m exactly?
Is this a sensible question?  Does it have a sensible answer?


Now produce a few different scattergraphs, maybe logarithmically
transformed.


## Task 2: econometric dataset

Download the file `e3mg.txt` from AUTonline and load it using


`e3mg <- read.table(file.choose(),header=TRUE)`

The first five columns are economic variables which can be controlled
politically [at least, some economists believe so].  The values are
normalized to lie in a standardised range.  Columns `rec_dep` and
`rec_len` are the length and depth of the resulting recession.  We
will use linear regression to predict `rec_dep` as a function of the
five independent variables.

* Using idiom like `summary(lm(rec_dep~oil.price + direct.tax +
  interest.rate, data=e3mg)`, fit a regression model to all five
  independent variables.

* By adding and removing variables judiciously, find the best model
  you can.  Here "best" means the best combination of simplicity and
  accuracy.


## Task 3: portfolio

You should by now have a small dataset which you are analysing for
your portfolio.  Apply linear regression to it and discuss your
results.

## Task 4: swiss dataset

The `swiss` dataset is an R builtin.  Type `swiss` at the R
commandline to see it, and `?swiss` to see some documentation.  Try to
predict infant mortality as a function of the other variables.

(note: be careful with case sensitivity, eg. `Infant.Mortality` rather
than `infant.Mortality` or `infant.mortality`).
