---
output: pdf_document
title: Linear regression
fontsize: 12pt
---


# introduction

This short handout takes you through linear regression.  Go through
the material, typing the R commands at the R prompt, making sure you
understand the content.
       
Remember that you can type `?lm` at the R prompt to get help
on that command.


# Linear regression: trees.

The R system includes a dataset called ```trees```:

```{r}
data(trees)
head(trees)
```

We will use linear regression to analyse this dataset.  We will start
by analysing the dependency of volume on tree height; The first step
is to plot a scattergraph:


```{r}
plot(Volume~Height,data=trees)
```

The points are reasonably scattered, without any outliers or
nonlinearity.  The points are showing broadly uniform variance about a
line.  We can use R to calculate the best-fit line using the
`lm()` function:


```{r}
lm(Volume~Height,data=trees)
```

This shows that the relation is approximately

\[
V =  -87.124    +    1.543H
\]

We can get more information by using the `summary()` command:


```{r}
fit <- lm(Volume~Height,data=trees)
summary(fit)
```

This gives more information such as p-values and standard errors. For
example, the slope is 1.5433 (estimated) but with an error of $\pm
0.3839$.


We can plot the regression line on the scattergraph as follows.

```{r}
plot(Volume~Height,data=trees)
abline(fit)
```

However, this model has a serious defect in that short trees are
predicted to have negative volumes, an absurdity.  We can mitigate
this defect by using a logarithmic transform:


```{r}
plot(Volume~Height, data=trees, log='xy')
```

The `plot()` command takes a `log` argument, in this case
instructing R to use logarithmic transforms on the x and y axes.

However, `abline()` does not work with this type of plot, so we will have to work manually:


```{r}
plot(log(Volume)~log(Height),data=trees)
logfit <- lm(log(Volume)~log(Height),data=trees)
abline(logfit)
```

And to see the details we use `summary()`:

```{r}
summary(logfit)
```


## Using Girth instead

The girth of the tree is a more useful predictor than height:

```{r}
plot(Volume~Girth,data=trees,log='xy')
```

We can try linear regression:


```{r}
logfit2 <- lm(log(Volume)~log(Girth),data=trees)
summary(logfit2)
```

The plot is more convincing than the numbers, IMHO:

```{r}
plot(log(Volume)~log(Girth),data=trees)
abline(logfit2)
```


It is possible to use both height and weight together as regressors:

```{r}
summary(lm(Volume ~ Girth + Height, data=trees))
```

This would mean that $V = -57.9877 + 4.7082G + 0.3393H$.  Again the
model has a defect that small trees are predicted to have negative
volume.  So it might be better to use logarithmic transformation:

```{r}
summary(lm(log(Volume) ~ log(Girth) + log(Height), data=trees))
```

This would mean that $\log V = -6.63 + 1.98\log G + 1.117\log H$,
or $V = e^{-6.63}G^{1.98}H^{1.117}$.

If the height and weight have combined effects we can use the star operator:


```{r}
summary(lm(log(Volume) ~ log(Girth) * log(Height), data=trees))
```

This would indicate that $\log V = -3.68 + 0.79\log G + 0.43\log H +
0.27\log G*\log H$.  Do you think this is a better model?


# Another dataset.


Download file `heightweight.txt` from AUTonline and load it into R, using

`a <- read.table(file.choose())`

Then use commands such as 


`plot(weight~height,data=a)`

to produce scattergraphs (one is shown below) but produce other ones,
maybe logarithmically transformed).

```{r}
a <- read.table("heightweight.txt")
plot(weight~height, data=a)
summary(lm(weight~height, data=a))
```

## your task

See how the predicted line is $w=60.138 + 11.343h$ from the R output
on the figure.  What is the predicted weight of someone
with a height of 2m exactly?  What is the predicted weight of someone
with a height of 0m exactly?  Is this sensible?

1. Using R idiom such as
`plot(weight~height,data=a,xlim=c(0,2.1),ylim=c(0,90))`, plot the
data on axes that include the origin; fit a line that goes through the
origin using `lm(weight~height+0,data=a)`.

2. Plot a straight line using `abline()` and observe that it does
not give a close fit.


Now consider that a logarithmic transformation may be used; see
the figure.


```{r}
plot(log(weight)~log(height), data=a)
fit <- lm(log(weight)~log(height), data=a)
summary(fit)
abline(fit)
```

Now the relationship is

\[
\log(w) = 4.24865 + 0.23994\log(h)
\]

where the numbers are from the R output.  This is equivalent to

\[
w = \exp(4.24865) h^{0.23994}
\]

or even (approximately).

\[
w = 70 h^{0.24}
\]

It is possible to regress against age using the `+` symbol:

```{r}
 summary(lm(weight~height+age,data=a))
```


but the high (and thus non-significant) p-value of `age` (0.266)
shows that the effect of age is not significant in this model.



