---
output: html_document
title: Fisher exact test suggested analysis
fontsize: 12pt
---

Suggested analysis follows.

# Example 1

```{r}
a <- read.table("handedness.txt",header=TRUE)
contingency <- table(a$gender,a$handedness)
dimnames(contingency) <- list(gender=c("F","M"),handedness=c("L","R"))
contingency
```


* What is the probability of a person in this dataset being Female? 16/21

  What is the probability of a person being Left-handed?  9/21

  What is the probability of a left handed person being male?  5/9

  What is the probability of a female being right-handed? 11/16


* State a sensible null hypothesis.  

  probability of a male being LH is equal to the probability of a
  female being LH; $p(LH|M) = p(LH|F)$.

* Translate `contingency` into the white ball, black ball language
  used in the R help page under `?hypergeometric`
  
  male -> white 
  
  female -> black
  
  LH -> drawn 
  
  RH -> left in bag
  
  
* Use `dhyper()` to calculate the probability of observing
  `contingency` with given marginal totals.

```{r}
 dhyper(4,5,16,9)
 ```

* Write out all possible contingency tables with the same marginal
  totals.  That is, all two-by-two tables with non-negative integer
  entries that have the same rowsums and columnsums as table
  `contingency` in the R code above.
  
```{r}
f <- function(n){matrix(c(9-n,n,7+n,5-n),2,2)}
L <- sapply(0:5,f,simplify=FALSE)
L
```
  
* Define "more extreme" to mean "contingency table with a smaller
  `dhyper()` probability", state the precise definition of p-value,
  and calculate it.  Is this a one-sided test or a two-sided test?

p-value is the probability, if the null is true, of obtaining the
observation or an observation more extreme.

```{r}	
(p <- dhyper(0:5,5,16,9))
sum(p)  #why do this?
sum(p[p<=p[5]])  # contains a slight infelicity, can you spot the issue?
```

* Verify with `fisher.test()`.

```{r}
fisher.test(contingency)
```


\newpage


# Example 2


```{r}
M <- matrix(c(7,1,2,5),2,2)
dimnames(M) <- list(professor=c(T,F),owns_car=c(T,F))
M
```

This shows incidence of car ownership among AUT academic staff.  One
might expect a professor, earning a massive salary, to be more likely
to be able to afford a car than a mere Senior Lecturer earning just
above minimum wage.

## your tasks

* State a sensible null hypothesis and specify whether a one-sided or
  two-sided test is more appropriate.

	Null: probability of owning a car is independent of professorial
    status.  One-sided test, more wealth implies higher prob of having
    a car.

* Use `fisher.test()` to test your null and interpret.


professor -> white

SL -> black

drawn -> car

not drawn -> no car

```{r}
M
fisher.test(M,alternative="greater")
```


* Verify using `dhyper()`.


```{r}
sum(dhyper(7:8,9,6,8))
```

(it is very easy to get this wrong:  one possible answer is `sum(dhyper(7:9,9,6,8))`
which gives the same answer because `dhyper(9,9,6,8)` is zero.  Enumerate the
sample space by writing out all possible tables and see if you can discover why 
`dhyper(9,9,6,8)==0`).


\newpage

# Example 3

Nine English and seven Scottish cows are tested for a particular
disease and six cows, of which four are Scottish, test positive.

## Your tasks

* Translate this into a contingency table
* State a sensible null and decide whether one-sided or two-sided tests are appropriate
* Test your null using `fisher.test()` and interpret.


```{r}
cows <- matrix(c(2,4,7,3),2,2)
dimnames(cows) <- list(nationality=c("English","Scottish"),disease=c(T,F))
cows
fisher.test(cows)
```

(two-sided, not significant)




# Example 4

Technically, Fisher's test applies only to two-by-two contingency
tables but is easily generalized to larger tables.  Enumerating the
sample space (that is, tables with matching marginal totals) is harder
than the two-by-two case, but easy with computers.  The null
distribution is not hypergeometric but again easy for computers to
evaluate.  In R, function `fisher.test()` takes contingency tables of
any size.  


The following table presents data on students' favourite
colours:


```{r}
M <- matrix(c(5,8,3,2,2,6,1,2,8),3,3)
dimnames(M) <- list(paper=c("ENGE502","STAT601","STAT805"),colour=c("red","blue","green"))
M
```

The following table presents data on a particular family in Auckland,
and how often they cooked different types of food over the last few
weeks.


```{r}
M <- matrix(c(1, 3, 5, 3, 6, 7, 5, 2, 2, 4, 2, 6, 2, 12, 3, 9, 4, 6, 0, 5),5,4)
dimnames(M) <- list(person=c("robin","lesley","alice","zac","annabel"),colour=c("pasta","soup","pizza","salad"))
M
```



## Your tasks

state sensible nulls for these two datasets and analyse using Fisher's
exact test.  Hint: you may need to use argument `simulate.p.value` to
enable R to operate (why do you think this is needed?)




