---
output: pdf_document
title: logistic regression
fontsize: 12pt
---

# Logistic regression


Consider the following dataset, taken from the discipline of human
psychology, in which a person is asked to identify whether or not
there is a particular feature in an image that is presented to the
subject.

```{r}
x <- seq(from=0,to=6,len=30)
y <- c(0,0,0,0,1,1,0,0,0,0,0,0,0,0,0,0,1,1,1,1,
          0,0,1,1,1,1,1,1,1,1)
```

Dataset y is the time in seconds that the subject is allowed to
view the image, and ``y'' is the result, with 0 meaning ``incorrect''
and 1 meaning ``correct''.

We can plot this easily:


```{r}
plot(y~x,pch=16)
```



# Doing logistic regression by hand

We can attempt to find the logistic fit by hand.  First we need to
define a logistic transformation:

```{r}
logistic <- function(x){exp(x)/(1+exp(x))}
LO <- function(p){log(p/(1-p))}
```


```{r}
p <- runif(100)
max(abs(p-logistic(LO(p))))
```

and

```{r}
z <- rnorm(100)
max(abs(z-LO(logistic(z))))
```

Make sure you understand what these lines do, and why we need both of them.
We can now use these functions to plot the regression line.


First of all we need a likelihood function.

```{r}
like <- function(beta){
  p <- logistic(beta[1] + beta[2]*x)
  success <- y==1
  fail    <- y==0
  prod(p[success]) * prod(1-p[fail])
}
```

so we can evaluate this at a couple of points:

```{r}
like(c(1,1))
like(c(1,2))
```

and we can use optim() to find the MLE:


```{r}
f <- function(beta){-log(like(beta))}
optim(c(1,1),f)
```

(a more mathematical approach would show that we do not need to worry
about local maxima).

# Logistic regression using glm() 

The regression is carried out usig the glm() function:



```{r}
fit <- glm(y~x,family='binomial')
fit
```

And we can see more detail using summary():


```{r}
summary(fit)
```

So, from the pvalue of 0.00425 we can see that the relationship is significant.
We can plot the regression using the logistic transform function.
Taking the coefficients from the fit, we have:
    
```{r}
plot(y~x,pch=16)
points(x,logistic(-3.07 + 0.95*x),type='l')
```



And this would show that the estimate for the value of x that gives $p=0.5$, that is $\hat{x} = -B/A = 3.07/0.95=3.23$, is about right.
