---
output: html_document
title: Pearson's chi square test
fontsize: 12pt
---

# Introduction

This short handout takes you through the basics of likelihood as
applied to the Gaussian distribution and the binomial distribution.
Go through the material, typing the R commands at the R prompt, making
sure you understand the content.
       
This short handout takes you through some theory and practical work
with the Poisson distribution; and shows how the Pearson Chi-square
test works.
  
Remember that you can type `?rbeta` at the R prompt to get help on the
beta distribution.  You are also encouraged to search for "Poisson
distributuion" and "Pearson chi square test" on Google; wikipedia is
an excellent source of information.


# Pearsons chi-square test
             
             
Pearsons chi-square test applies to a fixed, large, number $n$ of
trials, each one classified into one of $r$ categories.  I always
think of throwing $n$ balls into one of $r$ boxes and counting how
many balls end up in each of the boxes.

Suppose we have $n=100$ balls and $r=4$ boxes.  We can
simulate this situation as follows:
    
```{r}
boxes <- paste("box",1:4,sep="")
sample(boxes,100,replace=TRUE)
```

but of course it is much better to use `table()` to summarise:

```{r}
table(sample(boxes,100,replace=TRUE))
table(sample(boxes,100,replace=TRUE))
table(sample(boxes,100,replace=TRUE))
```
             
Try this yourself and get a feel for the variability.  Remember to
type `?sample` to get help on the the `sample()` command.

Suppose we do the experiment and observe

```{r}
o <- c(41,17,20,22)   # 'o' for 'observation'
```

Is this dataset consistent with a null hypothesis $H_0\colon
p_1=p_2=p_3=p_4=\frac{1}{4}$?  [what would such a null mean in
English?]  See how the observations are different from the expectation
of equal numbers in each box.  If the null is true, can chance
variability account for this observation?  In this case the expected
number of balls is 25 as the probabilities for the four boxes are
equal.

```{r}
e <- c(25,25,25,25)
```


We will assess our null using Pearsons chi square test.  You will
recall from lectures that it is possible to quantify the
goodness-of-fit for a distribution by calculating $B$:

$$
B=\sum_{i=1}^n\frac{\left(o_i-e_i\right)^2}{e_i}
$$

There is ready R idiom for calculating this:

```{r}
B <- sum((o-e)^2/e)  
B
```
Note the placement of the brackets (what does `sum(o-e)^2/e` give?).
We recall that, if the null is true, badness B is distributed with a
chi-square distribution with 3 degrees of freedom.  We can plot a
diagram, shown below.
    
    
```{r}
x <- seq(from=0,to=20,len=400)
plot(x,dchisq(x,df=3),type="l")
abline(h=0)
abline(v=B,lwd=6)
text(B,0.1,"observed B",pos=4)
```

Our observation is the value of B, which was 14.16 and this is shown
on the graph.  The p-value [being the probability, if the null is
true, of obtaining the observation or an observatin more extreme]
would be the area to the right of the line.  To calculate the exact
pvalue we use `pchisq()`:
    
```{r}
pchisq(14.16,df=3,lower.tail=FALSE)
```

which indicates that we may reject the null.  It is possible to test
other hypotheses.  Suppose that two of the boxes were twice as large
as the others and our hypothesis would be
$p_1=\frac{1}{3},p_2=\frac{1}{3}, p_3=\frac{1}{6}, p_4=\frac{1}{6}$.
Sampling from this null requires a new argument to sample():

```{r}
args(sample)
sample(boxes,100,replace=TRUE,prob=c(1/3,1/3,1/6,1/6))
table(sample(boxes,100,replace=TRUE,prob=c(1/3,1/3,1/6,1/6)))
table(sample(boxes,100,replace=TRUE,prob=c(1/3,1/3,1/6,1/6)))
table(sample(boxes,100,replace=TRUE,prob=c(1/3,1/3,1/6,1/6)))
```

[try your own probabilities here to get a feel for the variability;
again see how I build up the command from the inside out]


```{r}
probs <- c(1/3,1/3,1/6,1/6)
e <- probs*100
o <- c(29,31,20,20)  
B <- sum((o-e)^2/e)
pchisq(B,df=3,lower.tail=FALSE)
```

not significant because the pvalue exceeds 0.05.  Is this a type I
error, or a type II error, or something else?

# Random simulation

It is possible to verify that B is in fact approximately a chi-square
random variable by random sampling.  The overall process we have been
using is:

```{r}
n <- 100
p <- c(1/2,1/3,1/12,1/12)
e <- p*n
o <- tabulate(sample(1:4,n,replace=TRUE,prob=p))
B <- sum((o-e)^2/e)
B
```

We can create a histogram in one step.  See if you can unpick the
following R idiom:
    
 
```{r}
badness <- replicate(20000,sum((tabulate(sample(1:4,n,replace=TRUE,prob=p),nbins=4)-e)^2/e))
hist(badness)
```

Compare this with results from the `rchisq()` function in R.

# Pearsons chi-square test: horse kicks

We will assess a null hypothesis that the number of horse kick deaths
was distributed as a Poisson.  You will recall from lectures that it
is possible to quantify the goodness-of-fit for a distribution by
calculating $B$.  The Horse kick data was as follows:

```{r}
o <- c(109,65,22,3,1,0,0)
```

indicating that 109 regiments had zero deaths, 65 had 1, 22 had 2 and
so on.   We need to calculate deaths per regiment:
    
    
```{r}
regiments <- sum(o)
deaths    <- sum(o*(0:6))
deaths/regiments
```


We will combine the 3,1,0,0 together to form 4, giving data:
        
```{r}
o <- c(109,65,22,4)
```

So the ``4'' means 4 regiments had $\geqslant 3$ deaths.  We need to
calculate the probabilities of having $0,1,2,\geqslant 3$ deaths:
    
    
```{r}
dpois(0:2,lambda=0.61)
1-sum(dpois(0:2,lambda=0.61))
```
    
So the probabilities are:

```{r}
p <- c(dpois(0:2,lambda=0.61),1-sum(dpois(0:2,lambda=0.61)))
p
sum(p)
```

And the expected number of regiments are then:
    
```{r}
e <- 200*p
e
```


And we can calculate the badness:
    
```{r}
B <- sum((o-e)^2/e)
B
```

Now, if the null is true, B is distributed as a chi-square
distribution.  Look at the wikipedia page.  The difficult part of
Pearson's test is figuring out $d$, the number of degrees of freedom.
I'll cover this in lectures more fully, but for now we will just use
$d=2$.  So, if the null is true, we have $B\sim\chi^2_2$; see the
figure.  Recall the definition of p-value: the probability, if the
null is true, of obtaining the observation or one more extreme.  Here
the observation is B, the badness-of-fit, of 0.324.

```{r}
pchisq(0.324,df=2)
1-pchisq(0.324,df=2)
```

The second figure is the pvalue because `pchisq()}` gives the area to
the left and we want the area to the right.  This exceeds 5\% so is
not significant.


```{r}
hist(rchisq(1e6,df=2),nclass=100,col='gray')
abline(v=6,lwd=3)
```

The figure shows a histogram of the $\chi^2_2$ distribution.  A value
of B of 6 is shown; for this the p-value would be the area to the
right of the line.



# Your tasks


### task 1

Generate poisson observations using `rpois()`:

```{r}
set.seed(1266402)   # my student ID; use yours
table(rpois(200,lambda=0.61))
```


Go through the chi-square process above but with your randomly
generated dataset.  Observe that the null hypothesis is true for this
dataset.  Calculate B for your dataset and give a p-value as above.




